---
title: "data preparation"
author: "Sophie Moreau"
date: "2024-07-05"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(here)
library(readr)
library(viridis)

```

#Introduction
This is the supplementary information of the Master thesis. This document
covers the whole analysis performed to assess response diversity between fish communities
in Swiss peri-alpine lakes. 
This document has two parts: Data preparation (1) and Data analysis (2). 
Function to model general additive models is provided separately (see function_GAM.R). 
Code for figures used in the method figure is also provided separately (see supplementary_method_figure.R) 

```{r}
# do we want to include zurich and constance separately?
projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")

str(projet_lac_data)
projet_lac_data$LakeBasin <- as.factor(projet_lac_data$LakeBasin)

projet_lac_data |> 
  distinct(LakeBasin)
```

#1. Data preparation
##Abundance data
```{r}
##read in the data: Simstrat model data and Projet Lac combined
projet_lac_data <- read.csv("PLDB_final_short_depth_climate_cleantaxa_20052022.csv")

str(projet_lac_data)

# preparing additional column with abundance data of each species
abundance_data <- projet_lac_data |> 
  # exclude rows without taxa information
  filter(!Taxa_latin_FINAL %in% "NO_FISH") |> 
  # exclude samples obtained by electro-fishing
  filter(!Protocol %in% "electro") |> 
  #select columns relevant for species individual occurrences
 group_by(Lake, Basin3, Fishec_action, mean_temp_per_day, mean_last_7days, Weightg_raw, Weightg_soak, Depth_sample, Taxa_latin_FINAL, Protocol) |> 
  # prepare column with abundance of each species
  tally() |> 
  # values_fill = 0 added to make all NAs into 0
  pivot_wider(names_from = Taxa_latin_FINAL, values_from = n, values_fill = 0) 

# abundance of each species
abundance_data_long <- abundance_data |> 
  pivot_longer(cols = !(Lake : Protocol), names_to = "Species", values_to = "Abundance")

# add presence-absence column
abundance_data_long$Presence <- ifelse(abundance_data_long$Abundance >= 1, 1, 0)

# View(abundance_data_long)

max(abundance_data_long$Presence) #check if presence-absence only 0 and 1
max(abundance_data_long$Abundance)

#we only need information on species present in the lake, not all of them
abundance_data_long_lakepresence <- abundance_data_long |> 
  group_by(Lake, Species) |>
  # . stands for data being piped in do function
  do(LakePresence = max(.$Presence)) |> 
  unnest(cols = LakePresence)

# double-check 
abundance_data_long_lakepresence |>
  filter(LakePresence == "1") |> 
  group_by(Lake, LakePresence) |>
  count()

# join the two dfs
df_final <- abundance_data_long_lakepresence |> 
  filter(LakePresence == "1") |> 
  left_join(abundance_data_long )

#double-check if all lakepresence = 1
max(df_final$LakePresence)
min(df_final$LakePresence)

# save final df
saveRDS(df_final, "df_final.rds")

```

##Taxonomy and requirements
We pooled several species into ecomorphs to prevent loss of data-points. 
We also excluded species that were caught less than 10 times in the whole Projet Lac. 

```{r}
# read in prepared df
df_final <- readRDS("df_final.rds")

#######refining errors in species' identity that were noticed upon examination of the
# data frame, following the table 16 in the Projet Lac "Synthesis report"

# C. brienzii should not be present in lake Thun, add to Coregonus sp.

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii") & Lake %in% c("Thun"),
                          "Coregonus_sp", as.character(Species)))

# Salvelinus umbla in Lake Thun should be Salvelinus sp. 

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Salvelinus_umbla") & Lake %in% c("Thun"),
                          "Salvelinus_sp", as.character(Species)))


# Squalius squalus in Biel and Neuchatel to Squalius cephalus

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Squalius_squalus") & Lake %in% c("Biel", 
                                                                           "Neuchatel"),
                          "Squalius_cephalus", as.character(Species)))


#######Pooling species to be able to include more samples
# Profundal Cottus gobio are pooled

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c(
    "Cottus_gobio_Profundal_Thun",
    "Cottus_gobio_Profundal_Lucerne", "Cottus_gobio_Profundal_Walen"
  ),
  "Cottus_sp_Profundal", as.character(Species)
  ))


# Salvelinus is pooled into "profundal" and "limnetic" habitats

# profundal

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c( "Salvelinus_sp_Profundal_dwarf_Thun", 
                                          "Salvelinus_sp_Profundal_dwarf_VWS", 
                                          "Salvelinus_sp_Profundal_extreme_Thun",
    "Salvelinus_sp_Profundal_Walen_I", "Salvelinus_sp_Profundal_Walen_II"),
  "Salvelinus_sp_Profundal", as.character(Species)
  ))


# limnetic 
df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Salvelinus_sp_Limnetic_Thun", "Salvelinus_sp_Limnetic_VWS"),
                          "Salvelinus_sp_Limnetic", as.character(Species)))


# Coregonus species are pooled to new groups based on ecomorphs,
# following De-Dayne et al. 2022

# Albeli
df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_albellus", "Coregonus_candidus", "Coregonus_confusus",
                                         "Coregonus_heglingus", "Coregonus_zugensis"),
                          "Coregonus_sp_albeli", as.character(Species)))
# Balchen
df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_alpinus", "Coregonus_arenicolus", "Coregonus_duplex",
                        "Coregonus_helveticus", "Coregonus_palaea"),
                          "Coregonus_sp_balchen", as.character(Species)))

# Felchen

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_brienzii", "Coregonus_fatioi", "Coregonus_intermundia",
                                         "Coregonus_litoralis", "Coregonus_macrophthalmus", "Coregonus_zuerichensis"),
                          "Coregonus_sp_felchen", as.character(Species)))
# Large pelagic

df_final <- df_final |>
  mutate(Species = ifelse(Species %in% c("Coregonus_acrinasus", "Coregonus_wartmanni"),
                          "Coregonus_sp_large_pelagic", as.character(Species)))

# Benthic profundal
df_final <- df_final |> 
  mutate(Species = ifelse(Species == "Coregonus_profundus",
                          "Coregonus_sp_benthic_profundal", as.character(Species)))
# Pelagic profundal
df_final <- df_final |> 
  mutate(Species = ifelse(Species == "Coregonus_nobilis",
                          "Coregonus_sp_pelagic_profundal", as.character(Species)))

# "Coregonus_sarnensis": could not be assigned to Albeli or Felchen

# check species list
species <- df_final |> 
  distinct(Species)

# check how many species have less than 10 overall observations: 30
df_final |> 
  group_by(Species) |> 
  summarize(tot_abu = sum(Presence)) |> 
  filter(tot_abu <= 10)

# exclusion of species with less than 10 overall observations
df_models <- df_final |> 
  group_by(Species) |> 
  mutate(tot_abu = sum(Presence)) |> 
  ungroup() |> 
  filter(tot_abu >= 10) 

# save the df of the refined species, df will be used for the models
saveRDS(df_models, "df_models.rds")

df_models$Species <- as.factor(df_models$Species)
levels(df_models$Species)

species <- df_models |> 
  group_by(Species) |> 
  summarize(tot_obs = sum(Abundance))

```
##Data frames for GAMs
We compiled four data frames for the four types of general additive models that 
were run: Species with 1) binomial dataoccurring in one lake,
2) abundance data occurring in one lake, 3) binomial data 
occuring in multiple lakes, 4) abundance data and occuring in multiple lakes. 

```{r}
###Overview
#1. Separation of species into binomial and abundance data: sample with abundance > 1, 0 and 1 only -> binomial data
#2. species that occur only in one lake: Lake not added as random effect

#1. binomial vs. abundance data

df_models$Species <- as.character(df_models$Species)

non_binomial_species <- df_models |> 
  filter(Abundance > 1) |> 
  group_by(Species) |>
  distinct(Species) |> 
  pull(Species)


binomial_species <- df_models |> 
  filter(!Species %in% non_binomial_species) |> 
  distinct(Species) |> 
  pull(Species)


#2. binomial vs. abundance species occuring only in one lake

#1: GAM without lake as random effect, binomial

bi_one_occurence <- df_models |> 
  filter(Species %in% binomial_species) |>
  group_by(Species) |>
  summarize(n_lake = n_distinct(Lake)) |> 
  filter(n_lake == 1) |>
  pull(Species)

# save df
df_bi <- df_models |>
  filter(Species %in% bi_one_occurence)

saveRDS(df_bi, "data_frame_models/df_binomial_gam")

#2 GAM abundance without lake as random effect
abu_one_occurence <- df_models |> 
  filter(Species %in% non_binomial_species) |> 
  group_by(Species) |> 
  summarize(n_lake = n_distinct(Lake)) |> 
  filter(n_lake == 1) |>
  pull(Species)


df_abu <- df_models |> 
  filter(Species %in% abu_one_occurence)

saveRDS(df_abu, "data_frame_models/df_abundance_gam")

#binomial vs. abundance species occuring in several lakes

#3 gam with random intercept, binomial
bi_multi_occurence <- df_models |> 
  filter(Species %in% binomial_species) |> 
  group_by(Species) |> 
  summarize(Lakes = n_distinct(Lake)) |> 
  filter(!Lakes == 1) |> 
  pull(Species)

df_binomial_re <- df_models |> 
  filter(Species %in% bi_multi_occurence)

saveRDS(df_binomial_re, "data_frame_models/df_binomial_re")

#4 abundance GAM with random intercept
abu_multi_occurence <- df_models |> 
  filter(Species %in% non_binomial_species) |> 
  group_by(Species) |> 
  summarize(Lakes = n_distinct(Lake)) |> 
  filter(!Lakes == 1) |> 
  pull(Species)

df_abundance_re <- df_models |> 
  filter(Species %in% abu_multi_occurence)

saveRDS(df_abundance_re, "data_frame_models/df_abundance_re")

```