---
title: "drivers response diversity"
author: "Sophie Moreau"
date: "2024-07-05"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
# packages relevant for data analysis
library(tidyverse)
library(here)
library(readr)
library(readxl)
library(gratia)
library(mgcv)
library(forcats)
library(ggpubr)
library(RColorBrewer)
library(broom)
library(grid)
library(DHARMa)
library(mgcViz)
library(gridExtra)
library(grid)
library(viridis)
```

##Drivers of thermal response diversity

Preparing a df with the information on past eutrophication, species richness, and species categorization in each lake.
```{r}
# Df with properties of lakes
lake_info <- read_xlsx("lake_info.xlsx")

# select the columns we use
lake_selection <- lake_info |> 
  select(Lake, Phos_max, Max_depth, Lake_area)

# information about categorization in each lake for the subgroups
species_endemism <- read_excel("species_category_per_lake.xlsx") |> 
  rename(endemism = detail_category)

# exclude lepomis gibbosus and re-calculate species richness per lake
species_endemism_short <- species_endemism |> 
  filter(species != "Lepomis_gibbosus") |> 
  select(-sum_species) |> 
  group_by(fLake) |> 
  mutate(sum_species = sum(num_species))

species_endemism_short$endemism <- as.factor(species_endemism_short$endemism)

str(species_endemism_short)

# adding column with number of species per category in each lake
df_species_endemism <- species_endemism_short |> 
  select(-species, - num_species) |> 
  add_count(fLake, endemism) |> 
  rename(count = n) |> 
  distinct(fLake, count, sum_species, endemism)

str(df_species_endemism)

df_species_endemism$count <- as.numeric(df_species_endemism$count)

# making it wide 
df_species_endemism_long <- df_species_endemism |> 
  pivot_wider(names_from = endemism, values_from = count) |> 
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) |> 
  rename(Lake = fLake)

str(df_species_endemism_long)

# merge endemism info and lake info 

endemism_lake <- merge(df_species_endemism_long, lake_selection)

# Read in data on response diversity metrics in each lake
resp_div_all <- readRDS("total_models/resp_div_all.rds") |> 
  select(temp:fLake)

# select response diversity columns we need
rdiv <- resp_div_all |>  
  select(temp, fLake, rdiv, sign) |>
  drop_na()

# calculate mean and maximum values and rename Lake
df_rdiv <- rdiv |> 
  group_by(fLake) |> 
  mutate(mean_rdiv = mean(rdiv)) |> 
  mutate(max_rdiv = max(rdiv)) |> 
  mutate(mean_sign = mean(sign)) |> 
  mutate(max_sign = max(sign)) |> 
  distinct(fLake, mean_rdiv, max_rdiv, mean_sign, max_sign) |> 
  rename(Lake = fLake)

# combine lake properties, species subgroups and response diversity 

df_lm <- merge(df_rdiv, endemism_lake)

str(df_lm)

# phosphorus values as numeric
df_lm$Phos_max <- as.numeric(df_lm$Phos_max)

```

###Species richness and eutrophication
Multiple linear regression to calculate the effect of species richness and
eutrophication on response diversity.

###Fig. 3

```{r}
# Multiple linear regression for mean dissimilarity explained by species richness and past eutrophication: 
# lm(mean_rdiv  ~ species richness + eutrophication)
# Predict manually with new data eutrophication value fixed at mean and species richness min() to max()
# do the same the other way around: mean_rdiv  ~  eutrophication + species richness, with species richness fixed at mean and eutrophication min() to max()species richness +
#   prepare these two plots

# Phosphorus data on Poschiavo is not available
df_lm_excl <- df_lm |> 
  filter(Lake != "Poschiavo")


# first lm for species richness ~ maximum phosphorus 

lm1 <- lm(mean_rdiv ~ sum_species + Phos_max, data = df_lm_excl)

# checking model assumptions
summary(lm1)
shapiro.test(resid(lm1))
lmtest::bptest(lm1)
plot(lm1)

# model assumptions validated

# manually predict data
new_data <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))

# predict lm
prediction_lm1 <- predict.lm(lm1, newdata = new_data, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model1 <- cbind(new_data, prediction_lm1) |> 
  mutate(category ="color")



# second lm for maximum phosphorus ~ species richness
lm2 <- lm(mean_rdiv ~ Phos_max + sum_species, data = df_lm_excl)

# checking model assumptions
summary(lm2)
shapiro.test(resid(lm2))
lmtest::bptest(lm2)
plot(lm2)

# new data
new_data2 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))

# predict model
prediction_lm2 <- predict.lm(lm2, newdata = new_data2, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model2 <- cbind(new_data2, prediction_lm2) |> 
  mutate(category ="color")

# Plot Fig. 3
"#78BE63"
mycolor <- c("color" = "#78BE63")
# mycolor <- c("color" = "#A0E66A")
"98D47C"

lm1_plot <- bind_model1 |> 
  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  # ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

lm2_plot <- bind_model2 |> 
  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean dissimilarity") +
  xlab("Maximum total phosphorus (µg/L)") +
  # ylim(1,3) +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

# combine both plots
figure_3 <- ggarrange(lm1_plot, lm2_plot)

figure_3
# save as TIFF
# tiff(paste("total_models/plots/lm_species_richness_eutroph.tiff", sep = ""), units="in", width=12, height=5, res=300)
# 
# plot(figure_3)
# 
# # Closing the graphical device
# dev.off()

```

###Divergence figure 3

```{r}
# first lm for species richness ~ maximum phosphorus 

lm3 <- lm(mean_sign ~ sum_species + Phos_max, data = df_lm_excl)

# checking model assumptions
summary(lm3)
shapiro.test(resid(lm3))
lmtest::bptest(lm3)
plot(lm3)

# model assumptions validated

# manually predict data
new_data3 <- tibble(sum_species = seq(from = min(df_lm_excl$sum_species), to = max(df_lm_excl$sum_species), length = 50), Phos_max = mean(df_lm_excl$Phos_max))

# predict lm
prediction_lm3 <- predict.lm(lm3, newdata = new_data3, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model3 <- cbind(new_data3, prediction_lm3) |> 
  mutate(category ="color")



# second lm for maximum phosphorus ~ species richness
lm4 <- lm(mean_sign ~ Phos_max + sum_species, data = df_lm_excl)

# checking model assumptions
summary(lm4)
shapiro.test(resid(lm4))
lmtest::bptest(lm4)
plot(lm4)

# new data
new_data4 <- tibble(Phos_max = seq(from = min(df_lm_excl$Phos_max), to = max(df_lm_excl$Phos_max),length = 50), sum_species = mean(df_lm_excl$sum_species))

# predict model
prediction_lm4 <- predict.lm(lm4, newdata = new_data4, se.fit = TRUE, type = "response")

# combine new data and predictions
bind_model4 <- cbind(new_data4, prediction_lm4) |> 
  mutate(category ="color")

# Plot Fig. 3
mycolor <- c("color" = "#88D44F")
# mycolor <- c("color" = "#A0E66A")
"98D47C"

lm3_plot <- bind_model3 |> 
  ggplot(aes(sum_species, fit,  color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean divergence") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

lm4_plot <- bind_model4 |> 
  ggplot(aes(Phos_max, fit, color = category, fill = category)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)),  alpha = 0.3, color = NA) +
  theme_classic(base_size = 20) +
  ylab("Mean divergence") +
  xlab("Maximum total phosphorus (µg/L)") +
  scale_fill_manual(values = mycolor, aesthetics = c("color", "fill"), guide = NULL)

# combine both plots
figure_divergence <- ggarrange(lm3_plot, lm4_plot)

figure_divergence
# save as TIFF
# tiff(paste("total_models/plots/lm_divergence_multiple_linear.tiff", sep = ""), units="in", width=12, height=5, res=300)
# 
# plot(figure_divergence)
# 
# # Closing the graphical device
# dev.off()


```
###Community subgroups
####Fig. 4a
We fitted simple linear models for the number of species in each category and the mean dissimilarity in each lake. 
```{r}
# Lms for all categories: endemic, translcated, non_native and non_endemic_native
# with data on all lakes
# Endemic
lm_endemic <- lm(mean_rdiv ~ endemic, data = df_lm)

# checking model assumptions
summary(lm_endemic)
shapiro.test(resid(lm_endemic))
lmtest::bptest(lm_endemic)
plot(lm_endemic)

new_data_a <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
                                     length = 15))


prediction_endemic <- predict.lm(lm_endemic, newdata = new_data_a, se.fit = TRUE, type = "response")


df_lm_endemic <- cbind(new_data_a, prediction_endemic) |> 
  mutate(category = factor("endemic")) |> 
  rename(species_richness = endemic)

# Non-native

lm_nn <- lm(mean_rdiv ~ non_native, data = df_lm)

# checking model assumptions
summary(lm_nn)
shapiro.test(resid(lm_nn))
lmtest::bptest(lm_nn)
plot(lm_nn)

new_data_b <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
                                   length = 15))

prediction_nn <- predict.lm(lm_nn, newdata = new_data_b, se.fit = TRUE, type = "response")

df_lm_nn <- cbind(new_data_b, prediction_nn)  |> 
  mutate(category = factor("non_native")) |> 
  rename(species_richness = non_native)

# Non-endemic native

lm_nne <- lm(mean_rdiv ~ native, data = df_lm)

# checking model assumptions
summary(lm_nne)
shapiro.test(resid(lm_nne))
lmtest::bptest(lm_nne)
plot(lm_nne)


new_data_c <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
                                      length = 15))

prediction_nne <- predict.lm(lm_nne, newdata = new_data_c, se.fit = TRUE, type = "response")

df_lm_nne <- cbind(new_data_c, prediction_nne)  |> 
  mutate(category = factor("non_endemic_native")) |> 
  rename(species_richness = native)


# Translocated (non-native to region) species

lm_trans <- lm(mean_rdiv ~ non_native_region, data = df_lm)

# checking model assumptions
summary(lm_trans)
shapiro.test(resid(lm_trans))
lmtest::bptest(lm_trans)
plot(lm_trans)


new_data_d <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
                                              length = 15))

prediction_trans <- predict.lm(lm_trans, newdata = new_data_d, se.fit = TRUE, type = "response")

df_lm_trans <- cbind(new_data_d, prediction_trans)  |> 
  mutate(category = factor("translocated")) |> 
  rename(species_richness = non_native_region)


# Combine all four lm predictions
df_all_lms <- rbind(df_lm_endemic, df_lm_nn, df_lm_nne, df_lm_trans)


mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
               "translocated" = "#F2AD63")


plot_all_lms <- df_all_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category, linetype = category)) +  
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean dissimilarity") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"),
                    labels = c("endemic", "non-native (ns)", "non-endemic native", "translocated (ns)")) +
   scale_linetype_manual(values = c("endemic" = "solid",
                                   "non_native" = "dashed",
                                   "non_endemic_native" = "solid",
                                   "translocated" = "dashed"), labels = c("endemic", "non-native", "non-endemic native", "translocated"))

plot_all_lms
# save as TIFF
# tiff(paste("total_models/plots/lm_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(plot_all_lms)
# 
# dev.off()
```

####Divergence Fig. 4a

```{r}
# Lms for all categories: endemic, translcated, non_native and non_endemic_native
# with data on all lakes
# Endemic
deriv_lm1 <- lm(mean_sign ~ endemic, data = df_lm)

# checking model assumptions
summary(deriv_lm1)
shapiro.test(resid(deriv_lm1))
lmtest::bptest(deriv_lm1)
plot(deriv_lm1)

new_data_1 <- tibble(endemic = seq(from = min(df_lm$endemic), to = max(df_lm$endemic),
                                     length = 15))


prediction_endemic_1 <- predict.lm(deriv_lm1, newdata = new_data_1, se.fit = TRUE, type = "response")


df_deriv_endemic <- cbind(new_data_1, prediction_endemic_1) |> 
  mutate(category = factor("endemic")) |> 
  rename(species_richness = endemic)

# Non-native

deriv_lm2 <- lm(mean_sign ~ non_native, data = df_lm)

# checking model assumptions
summary(deriv_lm2)
shapiro.test(resid(deriv_lm2))
lmtest::bptest(deriv_lm2)
plot(deriv_lm2)

new_data_2 <- tibble(non_native = seq(from = min(df_lm$non_native), to = max(df_lm$non_native),
                                   length = 15))

prediction_2 <- predict.lm(deriv_lm2, newdata = new_data_2, se.fit = TRUE, type = "response")

df_deriv_nn <- cbind(new_data_2, prediction_2)  |> 
  mutate(category = factor("non_native")) |> 
  rename(species_richness = non_native)

# Non-endemic native

deriv_lm3 <- lm(mean_sign ~ native, data = df_lm)

# checking model assumptions
summary(deriv_lm3)
shapiro.test(resid(deriv_lm3))
lmtest::bptest(deriv_lm3)
plot(deriv_lm3)


new_data_3 <- tibble(native = seq(from = min(df_lm$native), to = max(df_lm$native),
                                      length = 15))

prediction_3 <- predict.lm(deriv_lm3, newdata = new_data_3, se.fit = TRUE, type = "response")

df_deriv_nne <- cbind(new_data_3, prediction_3)  |> 
  mutate(category = factor("non_endemic_native")) |> 
  rename(species_richness = native)


# Translocated (non-native to region) species

deriv_lm4 <- lm(mean_sign ~ non_native_region, data = df_lm)

# checking model assumptions
summary(deriv_lm4)
shapiro.test(resid(deriv_lm4))
lmtest::bptest(deriv_lm4)
plot(deriv_lm4)


new_data_4 <- tibble(non_native_region = seq(from = min(df_lm$non_native_region), to = max(df_lm$non_native_region),
                                              length = 15))

prediction_4 <- predict.lm(deriv_lm4, newdata = new_data_4, se.fit = TRUE, type = "response")

df_deriv_trans <- cbind(new_data_4, prediction_4)  |> 
  mutate(category = factor("translocated")) |> 
  rename(species_richness = non_native_region)


# Combine all four lm predictions
df_deriv_lms <- rbind(df_deriv_endemic, df_deriv_nn, df_deriv_nne, df_deriv_trans)


mycolors <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
               "translocated" = "#F2AD63")



lm_divergence_all <- df_deriv_lms|> 
  ggplot(aes(species_richness, fit, color = category, fill = category)) +
  geom_line(linewidth = 1, linetype = "dashed") +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), color = NA, alpha = 0.3) +
  theme_classic(base_size = 16) +
  ylab("Mean divergence") +
  xlab("Species richness") +
  scale_fill_manual(values = mycolors,
                    aesthetics = c("color", "fill"), 
                    labels = c("endemic", "non-native", "non-endemic native", "translocated"),
                    name = NULL)

lm_divergence_all

# save as TIFF
# tiff(paste("total_models/plots/lm_divergence_all.tiff", sep = ""), units="in", width=11, height=7, res=300)
# # plot(ggarrange(depth1, depth2, ncol = 2))
# # plot science discussion
# plot(lm_divergence_all)
# 
# dev.off()
```
#### Fig. 4b 
Predictions of all species shown inside the categories. 
```{r, fig.show='hide'}

# Plot species predictions inside categories across all lakes
# 5 categories;
# endemic: geographically constrained range
# non-native: from NA or Asia 
# non-endemic native: native to Switzerland but not endemic
# non-native region: native to Switzerland and surroundings but has been translocated
# to other Swiss lakes where the species was not native
# endemic translocated: endemic species that were translocated to other lakes

# read in table with the categories and rename the categories
species_category <- read_excel("species_category_across_lakes.xlsx") |> 
  select(-notes) |> 
  mutate(category = ifelse(category == "native", "non_endemic_native", category)) |> 
  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
  mutate(category2 = factor("all"))

species_category$category <- as.factor(species_category$category)
levels(species_category$category)

# merge predictions with categories
model_pred_categories <- merge(model_predictions, species_category) 

# change category names for facet_wrap
category_names <- c(
  `endemic` = "endemic",
  `endemic_and_translocated` = "translocated endemic",
  `non_endemic_native` = "non-endemic native",
  `non_endemic_native_and_translocated` = "translocated non-endemic native",
  `non_native` = "non-native"
)

# colors for the plot
mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")

# Rescale the predictions to 0-1 
model_pred_categories_rescaled <- model_pred_categories |> 
  group_by(species) |> 
  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))

# Plot the rescaled predictions
plot_pred <- model_pred_categories_rescaled |>
  # exclude Lepomis gibbosus
  filter(species != "Lepomis_gibbosus") |> 
  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
  geom_line(linewidth = 0.8) +
  theme_bw(base_size = 16) +
  ylab("Abundance") +
  xlab("Temperature") +
  scale_color_manual(values = mycolors1, guide = NULL) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "bottom",
        legend.title = element_blank()) 

plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))

plot_category_predictions

# save plot as TIFF
# tiff(paste("total_models/plots/plot_category_predictions.tiff", sep = ""), units="in", width=15, height=4, res=300)
# 
# plot(plot_category_predictions)
# 
# dev.off()

```






