---
title: "data preparation"
author: "Sophie Moreau, Conor Waldock"
date: "2024-07-05"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}

pacman::p_load(tidyverse, here, readr, viridis)

```

#Introduction
Data preparation for analysis.

```{r}

# read in data
projet_lac_data <- read.csv("data/PLDB_final_short_depth_climate_cleantaxa_20052022.csv")

str(projet_lac_data)

projet_lac_data$LakeBasin <- as.factor(projet_lac_data$LakeBasin)

projet_lac_data |> 
  distinct(LakeBasin)

```

#1. Data preparation
##Abundance data
```{r}

# preparing additional column with abundance data of each species
abundance_data <- projet_lac_data |> 
  # exclude rows without taxa information
#  filter(!Taxa_latin_FINAL %in% "NO_FISH") |>  # here I think this should be included
  # exclude samples obtained by electro-fishing
  filter(!Protocol %in% "electro") |> 
  #select columns relevant for species individual occurrences
  select(Lake, Basin3, Fishec_action, mean_temp_per_day, mean_last_7days, Num_indiv_soak, Depth_sample, Taxa_latin_FINAL, Protocol) |>
  # turn number into whole number of negative binomial models
  mutate(n = case_when(Num_indiv_soak < 1 ~ 1, T ~ round(Num_indiv_soak))) %>% 
  # sum abundances across fishing event
  group_by(Lake, Basin3, Fishec_action, mean_temp_per_day, mean_last_7days, Depth_sample, Taxa_latin_FINAL, Protocol) %>% 
  summarise(n = sum(n, na.rm = T)) %>% 
  # values_fill = 0 added to make all NAs into 0
  pivot_wider(names_from = Taxa_latin_FINAL, values_from = n, values_fill = 0)


# abundance of each species
abundance_data_long <- abundance_data |> 
  pivot_longer(cols = !(Lake : Protocol), names_to = "Species", values_to = "Abundance")

# add presence-absence column
abundance_data_long$Presence <- ifelse(abundance_data_long$Abundance >= 1, 1, 0)

# View(abundance_data_long)

max(abundance_data_long$Presence) #check if presence-absence only 0 and 1
max(abundance_data_long$Abundance)

# check plot of abundances

ggplot(data = abundance_data_long %>% filter(Abundance > 0)) + geom_point(aes(x = log10(Abundance+1), y = Species))

# we only need information on species present in the lake, not all of them
abundance_data_long_lakepresence <- abundance_data_long |> 
  group_by(Lake, Species) |>
  # . stands for data being piped in do function
  do(LakePresence = max(.$Presence)) |> 
  unnest(cols = LakePresence)

# double-check 
abundance_data_long_lakepresence |>
  filter(LakePresence == "1") |> 
  group_by(Lake, LakePresence) |>
  count() %>% 
  arrange(n)

# join the two dfs so that we filter the large dataset only to lakes where species are present in project Lac
# (e.g., not modelling absences in lakes with only absences)
df_final <- abundance_data_long_lakepresence |> 
  filter(LakePresence == "1") |> 
  left_join(abundance_data_long)

#double-check if all lakepresence = 1
max(df_final$LakePresence)
min(df_final$LakePresence)

# check number of occurrences per lake
df_final %>% 
  group_by(Species, Lake) %>% 
  filter(Presence == 1) %>% 
  count %>% 
  view

# Now remove "NO_FISH" as 0s are added in through the pivot table
df_final <- df_final %>% 
  filter(Species != 'NO_FISH')
  
```

## Checking impact of excluding electrofishing
```{r}
# Which fish are caught mostly with electrofishing?
projet_lac_data %>% 
  group_by(Taxa_latin_FINAL, Type_fishing ) %>% 
  count %>% 
  group_by(Taxa_latin_FINAL) %>% 
  filter(n==max(n)) %>% 
  filter(Type_fishing == 'electrofishing') %>% 
  pull(Taxa_latin_FINAL)

# This is important to know - could re-evaluate whether to include these species and use the fishing type as a factor in the modelling

```


##Taxonomy and requirements

We pooled several species into ecomorphs to prevent loss of data-points. 
We also excluded species that were caught less than 10 times in the whole Projet Lac. 

```{r}


# Here I now manually matched all these to our decided ecotypes for the species and corrected names in specific lakes based on updated taxonomic revisions. 
final_taxonomy <- read_csv('data/taxonomy_May2025/species_lake_list_raw_EDITED.csv')

# join in taxonomic fixes to data
df_final2 <- left_join(df_final, final_taxonomy %>% select(Species, Lake, Basin3, Species_ecotype, Swisswide_biogeographic_status, Lake_biogeographic_status) %>% unique())

# check species list
species <- df_final2 |> 
  distinct(Species_ecotype) %>% 
  pull(Species_ecotype) %>% 
  sort

species

# which are missing?
df_final2 %>% 
  filter(is.na(Species_ecotype)) %>% 
  pull(Basin3)

# check how many species have less than 10 overall observations: 30
df_final2 |> 
  group_by(Species_ecotype) |> 
  summarize(prevelance = sum(Presence, na.rm = T)) |> 
  filter(prevelance <= 10) %>% 
  arrange(prevelance) %>% 
  print(n = 100)

# Comment on each species being removed from the analysis. Key here is if the species provides unique information so should try 
# and keep even if few records. 
# Alosa_fallax    # Keep - Contributes uniquely to Maggiore with Alosa alosa - Keep
# Ameiurus_melas  # Keep - Contributes althought having few records to Geneva and Magiorre - Keep
# Barbus_plebejus # Remove - just one record - Contributes uniquely to Maggiore
# Carassius_gibelio     # Keep - Contributes uniquely to Magiorre and Lugano- Keep
# Chondrostoma_nasus    # Remove - But check again if using electrofishing
# Chondrostoma_soetta   # Remove - But check again if using electrofishing
# Coregonus_sp_pelagic_profundal # Deep water pelagic form from Lake lucern - Combine with profundal?
# Cottus_gobio_Rhine    # Keep - these are the only forms in Geneva and Constance 
# Esox_cisalpinus       # Keep - but might be better to combine as Esox spp. for both north and south of alps?
# Gasterosteus_gymnurus # Remove
# Gobio_obtusirostris   # Remove - only one presences in Constance - Danubian lineage - Gobio gobio also present - Remove
# Oncorhynchus_mykiss   # Keep - Non-native in Constance and Maggiore - Keep
# Padogobius_bonelli    # Remove - In Po lineage only - probably mostly caught with electrofishing.
# Phoxinus_septimaniae  # Remove - Only present in poschiavo and chalain
# Phoxinus_sp           # Remove as csikii in Pochiavo
# Rutilus_aula          # should remove as Pigus is not in the sourthern lakes we consider here so we cannot assign a southern Rutilus
# # df_final %>% filter(grepl('Rutilus', Species), Basin3 == 'Maggiore', Abundance != 0) %>% ggplot(data = .) + geom_histogram(aes(x = Depth_sample, col = Species))
# Salaria_fluviatilis_French # combined now into Salaria fluviatilis - still few records as most records are from electrofishing
# Salaria_fluviatilis_Italian # combined now into Salaria fluviatilis - still few records as most records are from electrofishing
# Salmo_cenerinus # only in poschiavo - cannot model with just 1 dataset
# Salmo_labrax    # only in Sils - 3 records so probably cannot be modelled reliably
# Salmo_marmoratus # In 3 lakes in the Po catchment but only 4 records..
# Salmo_sp
# Salvelinus_namaycush # Only recorded in Poschiavo
# Salvelinus_profundus # This gets put into profundal form across all lakes
# Salvelinus_sp           # These are unidentified species - not sure if to keep these
# Salvelinus_sp_Limnetic  # Should try and keep as it is a unique form
# Telestes_muticellus     # Mostly caught with electrofishing
# Thymallus_thymallus     # Mostly caught with electrofishing

# Manually remove those based on decisions above of species with less than 10 overall observations
# RE-EVALUATE THIS IF USING ELECTROFISHING SURVEYS
df_models <- df_final2 |> 
  filter(!Species_ecotype %in% c("Phoxinus_sp", 
                         "Barbus_plebejus", "Oncorhynchus_mykiss", "Salmo_sp",
                         "Chondrostoma_nasus", "Chondrostoma_soetta", "Gasterosteus_gymnurus", 
                         "Gobio_obtusirostris", "Padogobius_bonelli", "Phoxinus_septimaniae", 
                         "Rutilus_aula", "Salmo_cenerinus", "Salmo_labrax", "Salmo_marmoratus", 
                         "Telestes_muticellus", "Lampetra_planeri", "Rhodeus_amarus", 
                         "Alburnoides_bipunctatus", "Thymallus_thymallus"))
  
species <- df_models |> 
  group_by(Species_ecotype) |> 
  summarize(number_of_records = sum(Presence),
    total_abundance = sum(Abundance))

species

```

##Data frames for GAMs

We compiled four data frames for the four types of general additive models that were run: Species with: 
1) binomial data occurring in one lake, 
2) abundance data occurring in one lake, 
3) binomial data  occuring in multiple lakes,
4) abundance data and occuring in multiple lakes. 

```{r}


###Overview
#1. Separation of species into binomial and abundance data: sample with abundance > 1, 0 and 1 only -> binomial data
#2. species that occur only in one lake: Lake not added as random effect

df_models_final <- df_models %>% 
  # remove species original column
  select(-Species) %>% 
  # rename ecotype as species
  rename(Species = Species_ecotype) %>% 
  group_by(Species) %>% 
  mutate(model_type = case_when(max(Abundance) > 1 ~ 'zip', T ~ 'bin'),
         lake_random = case_when(n_distinct(Lake) > 1 ~ T, T ~ F)) %>% 
  filter(!is.na(Depth_sample))
  
# save the df of the refined species, df will be used for the models
saveRDS(df_models_final, "data/df_models_final.rds")

# checking logic worked as expected
df_models_final %>% 
  group_by(Species) %>% 
  mutate(max_abundance = max(Abundance), 
         n_lake = n_distinct(Lake)) %>% 
  select(n_lake, Species, max_abundance, lake_random, model_type) %>% 
  unique() %>% 
  print(n = 52)

``` 