---
title: "02_models_Electrofishing"
author: "Sophie Moreau, Conor Waldock"
date: "2024-09-27"
output: html_document
editor_options: 
  chunk_output_type: console
---
#General Additive Models
Running the predictions for all GAMs with the smaller data frames to reduce running 
time. 

```{r}
# packages relevant for data analysis
pacman::p_load(
  tidyverse,
  here,
  readr,
  readxl,
  gratia,
  mgcv,
  forcats,
  ggpubr,
  RColorBrewer,
  broom,
  grid,
  DHARMa,
  mgcViz,
  gridExtra,
  viridis, 
  modelsummary,
  gratia, 
  broom,
  tidyverse,
  DHARMa,
  mgcv,
  gratia,
  mgcViz
)

# function for the temperature only GAMs with all run as expected
source("scripts/functions/predictions_function.R")

```

# Here modify parameters and then run scripts below to produce the following outputs

1. Fit models and save all objects
2. Plots for gamchecks
3. Plot of response curves for all species
4. Summary tables of model for all species

```{r}

# LOAD TESTING PARAMETERS
model_run = 'run_01052025'

# LOAD DATA
df_models_final <- readRDS('data/df_models_final_ELECTRO.rds')

```


# Fit models across multiple k-values
```{r}

# x <- 'Rutilus_rutilus' # TEST CODE WITH ONE SPECIES

lapply(c(3,4,5), function(set_k){
  
  # save outputs to folder
  output_folder <- paste0('outputs/', model_run, '_electro', '_k=', set_k)
  dir.create(output_folder, recursive = T)
  
  # FIT ALL GAMS
  
  all_gams <- lapply(unique(df_models_final$Species), function(x){
  
  tryCatch({
  print(x)
  
  model_data <- df_models_final %>% 
    filter(Species == x) %>% 
    mutate(fProtocol = as.factor(Protocol), 
           fLake = as.factor(Lake))
  model_type <- unique(model_data$model_type)
  lake_random <- unique(model_data$lake_random)
  
  if(model_type == 'bin' & lake_random == F){
  # binomial model
    gam_output <- gam(data = model_data, Presence ~ s(mean_last_7days, k = set_k) +
                            s(fProtocol, bs = 're'), family = binomial)
  }
  if(model_type == 'zip'& lake_random == F){
  # zero-inflated poisson model
    gam_output <- gam(data = model_data, Abundance ~ s(mean_last_7days, k = set_k) +
                    s(fProtocol, bs = 're'), family = ziP())
  }
  if(model_type == 'bin' & lake_random == T){
  # binomial model
    gam_output <- gam(data = model_data, Presence ~ s(mean_last_7days, k = set_k) +
                            s(fProtocol, bs = 're') + s(fLake, bs = 're'), family = binomial)
  }
  if(model_type == 'zip'& lake_random == T){
  # zero-inflated poisson model
    gam_output <- gam(data = model_data, Abundance ~ s(mean_last_7days, k = set_k) +
                    s(fProtocol, bs = 're') + s(fLake, bs = 're'), family = ziP())
  }

  return(gam_output)
  
  }, error = function(e){NA})
  
})
  
  names(all_gams) <- unique(df_models_final$Species)

  # print which species failed
  is.na(all_gams) %>% sum
  writeLines(unique(df_models_final$Species)[which(is.na(all_gams))], file.path(output_folder, "missing_species.txt"))

  ## save all the gams
  saveRDS(all_gams[which(!is.na(all_gams))], paste0(output_folder, 'all_gams.RDS'))

  
  #### PLOT SIMULATION OUTPUT AS VALIDATION OF MODEL STRUCTURE

  gam_check_folder <- file.path(output_folder, 'gam_check')

  dir.create(gam_check_folder, recursive = T)

  lapply(names(all_gams), function(x){tryCatch({

        # prepare residuals
        simulationOutput <- simulateResiduals(fittedModel = all_gams[[x]], plot = F)
        tiff(paste(gam_check_folder, '/', x, ".tiff", sep = ""), width = 800, height = 600)
        print(plot(simulationOutput))
        dev.off()
        
        }, error = function(e) NA)
    
    })



  ##### GET PREDICTIONS AS STANDARIZED OBJECT FOR PLOTTING WITH GGPLOT2

  model_temp <- expand.grid(mean_last_7days = seq(
          from = min(df_models_final$mean_last_7days, na.rm = TRUE),
          to = max(df_models_final$mean_last_7days, na.rm = TRUE), by = 0.02),
          fProtocol = 'new_protocol', 
          fLake = 'new_lake')

  # get predictions across model table (model_temp)
  preds <- lapply(names(all_gams), function(x){tryCatch({
  
    pred_x <- predict(all_gams[[x]], model_temp, exclude = c('fProtocol', 'fLake'), type = "response", se = T)

  }, error = function(e) list(fit = NA, se.fit = NA))
    
    })

  # process outputs from lapply look
  all_preds <- data.frame(lapply(preds, function(x) x$fit))
  names(all_preds) <- names(all_gams)
  
  all_se <- data.frame(lapply(preds, function(x) x$se.fit))
  names(all_se) <- names(all_gams)
  
  # bind with predicted values
  all_preds_long <- bind_cols(model_temp, all_preds) %>% 
    pivot_longer(4:ncol(.))
  
  all_se_long <- bind_cols(model_temp, all_se) %>% 
    pivot_longer(4:ncol(.), values_to = 'se')
  
  # join together
  all_preds_long <- left_join(all_preds_long, all_se_long)

  
  # Calculate the realized limits of the species based on observations
  thermal_limits <- df_models_final %>% 
    filter(Presence == 1) %>% 
    group_by(Species) %>% 
    summarise(
      across(mean_last_7days, list(min = ~min(.x, na.rm = TRUE),
                                   max = ~max(.x, na.rm = TRUE)))
    ) %>% 
    rename(name = Species)
  
  # Join in limits
  all_preds_long_limits <- left_join(thermal_limits, all_preds_long)
  
  # Add logic to colour line depending on realized limits
  all_preds_long_limits <- all_preds_long_limits %>% 
    mutate(colour = case_when(mean_last_7days > mean_last_7days_min & mean_last_7days < mean_last_7days_max ~ 'inside', T ~ 'outside'))
  
  # Add identifier if predictions are for with or without electrofishing data
  all_preds_long_limits <- all_preds_long_limits %>% 
    mutate(including_electro = T)
  
  # Save output
    saveRDS(all_preds_long_limits %>% 
              mutate(k = set_k), 
            paste0(output_folder, '/temperature_predictions.RDS'))

  
  png(filename = paste0(output_folder, '/thermal_response_curves', '_k=', set_k, '.png'), res = 300, height = 3000, width = 4000)
  print(ggplot(data = all_preds_long_limits %>% 
                 filter(!grepl('Salvelinus_sp_Pro', name))) + 
    geom_line(aes(x = mean_last_7days, y = value, group = name, col = colour)) + 
    scale_colour_manual(values = c('red', 'gray50')) + 
    facet_wrap(~name, scales = 'free_y') + 
    theme_bw() + 
    theme(strip.background = element_blank(), 
          strip.text = element_text(size = 7)))
  dev.off()



  #### SAVE SUMMARISE OF MODEL PROPERTIES
  
  tidy(all_gams[[1]])
  
  tidy_output <- lapply(names(all_gams), function(x){tryCatch({
    
    tidy_x <- tidy(all_gams[[x]])
    tidy_x$dev.exp <- signif(summary(all_gams[[x]])$dev.expl[1],2)
    tidy_x$aic <- all_gams[[x]]$aic
    tidy_x$species_name <- x
    tidy_x
  }, error = function(e) NULL)
    
  })

  table_model_summaries <- bind_rows(tidy_output) %>% 
    relocate(species_name)
  
  write_csv(table_model_summaries, 
            file = paste0(output_folder, '/model_summaries.csv'))
  
  hist(unique(table_model_summaries$dev.exp))
  write_csv(table_model_summaries %>% 
                select(species_name, dev.exp) %>% 
                unique() %>% 
                arrange(dev.exp), file = paste0(output_folder, '/model_devexp.csv')) # really good model deviance explained

  hist(unique(table_model_summaries$aic))
  write_csv(table_model_summaries %>% 
                select(species_name, aic) %>% 
                unique() %>% 
                arrange(aic), file = paste0(output_folder, '/model_aic.csv')) # really good model deviance explained

  
#### END OF LOOPING ACROSS K VALUES

})

```


# read in k=3 for neater plot for supporting materials
```{r}
  # Save output
  all_preds_long_limits <- readRDS("outputs/run_01052025_electro_k=3/temperature_predictions.RDS")

  all_preds_long_limits <- 
    all_preds_long_limits %>% 
    mutate(name = gsub('_', ' ', name))
  
  png(filename = 'figures/thermal_response_curves__k=3.png', res = 300, height = 3000, width = 3000)
  print(ggplot(data = all_preds_long_limits %>% 
                 mutate(value = case_when(value > 500 ~ 500, T ~ value))) + 
    geom_line(aes(x = mean_last_7days, y = value, group = name, col = colour)) + 
    scale_colour_manual(name = 'inside sampling', values = c('red', 'gray50')) + 
    facet_wrap(~name, scales = 'free_y', ncol = 5) + 
    theme_bw() + 
    theme(strip.background = element_blank(), 
          strip.text = element_text(size = 7),
          legend.position = 'bottom'))+
    xlab('Temperature') + 
    ylab('Predicted abundance')
  dev.off()

```



# READ IN ALL DIFFERENT k VALUES AND CHECK THE RESPONSE CURVES ACROSS ALL SPECIES 
```{r}


# bind together no electro and electro
all_preds <- bind_rows(lapply(list.files('outputs', pattern = 'temperature_predictions.RDS', full.names = T, recursive = T), readRDS))

# plot together all response curves
png('figures/thermal_responses_compare_methods_k=3.png', res = 300, height = 3000, width = 4000)
ggplot(data = all_preds %>% 
         filter(colour == 'inside', k == 3)) + 
    geom_line(aes(x = mean_last_7days, y = value, col = including_electro)) + 
    scale_colour_discrete() + 
    facet_wrap(~name, scales = 'free_y') + 
    theme_bw() + 
    theme(strip.background = element_blank(), 
          strip.text = element_text(size = 7))
dev.off()

# plot together all response curves - no 10
png('figures/thermal_responses_together_no10_electro.png', res = 300, height = 3000, width = 4000)
ggplot(data = all_preds %>% 
         filter(colour == 'inside')) + 
    geom_line(aes(x = mean_last_7days, y = value, col = paste0(k), lty = including_electro)) + 
    scale_colour_discrete() + 
    facet_wrap(~name, scales = 'free_y') + 
    theme_bw() + 
    theme(strip.background = element_blank(), 
          strip.text = element_text(size = 7))
dev.off()


```









