---
title: "02_models_Electrofishing_Depth"
author: "Sophie Moreau, Conor Waldock"
date: "2024-09-27"
output: html_document
editor_options: 
  chunk_output_type: console
---
#General Additive Models
Running models but with depth as a covariate to test whether the thermal response curves are the same. 

```{r}
# packages relevant for data analysis
pacman::p_load(
  tidyverse,
  here,
  readr,
  readxl,
  gratia,
  mgcv,
  forcats,
  ggpubr,
  RColorBrewer,
  broom,
  grid,
  DHARMa,
  mgcViz,
  gridExtra,
  viridis, 
  modelsummary,
  gratia, 
  broom,
  tidyverse,
  DHARMa,
  mgcv,
  gratia,
  mgcViz
)

# function for the temperature only GAMs with all run as expected
source("scripts/functions/predictions_function.R")

```

# Here modify parameters and then run scripts below to produce the following outputs

1. Fit models and save all objects
2. Plots for gamchecks
3. Plot of response curves for all species
4. Summary tables of model for all species

```{r}

# LOAD TESTING PARAMETERS
model_run = 'run_01052025_depth'

# LOAD DATA
df_models_final <- readRDS('data/df_models_final_ELECTRO.rds')

```


# Fit models across multiple k-values
```{r}

# x <- 'Rutilus_rutilus' # TEST CODE WITH ONE SPECIES

lapply(c(3), function(set_k){
  
  # save outputs to folder
  output_folder <- paste0('outputs/', model_run, '_electro', '_k=', set_k)
  dir.create(output_folder, recursive = T)
  
  # FIT ALL GAMS
  
  all_gams <- lapply(unique(df_models_final$Species), function(x){
  
  tryCatch({
  print(x)
  
  model_data <- df_models_final %>% 
    filter(Species == x) %>% 
    mutate(fProtocol = as.factor(Protocol), 
           fLake = as.factor(Lake))
  model_type <- unique(model_data$model_type)
  lake_random <- unique(model_data$lake_random)
  
  if(model_type == 'bin' & lake_random == F){
  # binomial model
    gam_output <- gam(data = model_data, Presence ~ s(mean_last_7days, k = set_k) + s(Depth_sample, k = set_k) +
                            s(fProtocol, bs = 're'), family = binomial)
  }
  if(model_type == 'zip'& lake_random == F){
  # zero-inflated poisson model
    gam_output <- gam(data = model_data, Abundance ~ s(mean_last_7days, k = set_k) + s(Depth_sample, k = set_k) +
                    s(fProtocol, bs = 're'), family = ziP())
  }
  if(model_type == 'bin' & lake_random == T){
  # binomial model
    gam_output <- gam(data = model_data, Presence ~ s(mean_last_7days, k = set_k) + s(Depth_sample, k = set_k) +
                            s(fProtocol, bs = 're') + s(fLake, bs = 're'), family = binomial)
  }
  if(model_type == 'zip'& lake_random == T){
  # zero-inflated poisson model
    gam_output <- gam(data = model_data, Abundance ~ s(mean_last_7days, k = set_k) + s(Depth_sample, k = set_k) +
                    s(fProtocol, bs = 're') + s(fLake, bs = 're'), family = ziP())
  }

  return(gam_output)
  
  }, error = function(e){NA})
  
})
  
  names(all_gams) <- unique(df_models_final$Species)

  # print which species failed
  is.na(all_gams) %>% sum
  writeLines(unique(df_models_final$Species)[which(is.na(all_gams))], file.path(output_folder, "missing_species.txt"))

  ## save all the gams
  saveRDS(all_gams[which(!is.na(all_gams))], paste0(output_folder, 'all_gams.RDS'))
  
  #### PLOT SIMULATION OUTPUT AS VALIDATION OF MODEL STRUCTURE

  gam_check_folder <- file.path(output_folder, 'gam_check')

  dir.create(gam_check_folder, recursive = T)

  lapply(names(all_gams), function(x){tryCatch({

        # prepare residuals
        simulationOutput <- simulateResiduals(fittedModel = all_gams[[x]], plot = F)
        tiff(paste(gam_check_folder, '/', x, ".tiff", sep = ""), width = 800, height = 600)
        print(plot(simulationOutput))
        dev.off()
        
        }, error = function(e) NA)
    
    })


  ##### GET PREDICTIONS FOR TEMPERATURE

  model_temp <- expand.grid(mean_last_7days = seq(
    from = min(df_models_final$mean_last_7days, na.rm = TRUE),
    to = max(df_models_final$mean_last_7days, na.rm = TRUE), by = 0.02),
    fProtocol = 'new_protocol', 
    fLake = 'new_lake')

  # get predictions across model table (model_temp)
  preds <- lapply(names(all_gams), function(x){tryCatch({
    
    sp_depth = df_models_final %>% filter(Species == x, Presence == 1) %>% pull(Depth_sample) %>% mean
  
    pred_x <- predict(all_gams[[x]], data.frame(model_temp, Depth_sample = sp_depth), exclude = c('fProtocol', 'fLake'), type = "response", se = T)

  }, error = function(e) list(fit = NA, se.fit = NA))
    
    })

  # process outputs from lapply look
  all_preds <- data.frame(lapply(preds, function(x) x$fit))
  names(all_preds) <- names(all_gams)
  
  all_se <- data.frame(lapply(preds, function(x) x$se.fit))
  names(all_se) <- names(all_gams)
  
  # bind with predicted values
  all_preds_long <- bind_cols(model_temp, all_preds) %>% 
    pivot_longer(5:ncol(.))
  
  all_se_long <- bind_cols(model_temp, all_se) %>% 
    pivot_longer(5:ncol(.), values_to = 'se')
  
  # join together
  all_preds_long <- left_join(all_preds_long, all_se_long)

  
  # Calculate the realized limits of the species based on observations
  thermal_limits <- df_models_final %>% 
    filter(Presence == 1) %>% 
    group_by(Species) %>% 
    summarise(
      across(mean_last_7days, list(min = ~min(.x, na.rm = TRUE),
                                   max = ~max(.x, na.rm = TRUE)))
    ) %>% 
    rename(name = Species)
  
  # Join in limits
  all_preds_long_limits <- left_join(thermal_limits, all_preds_long)
  
  # Add logic to colour line depending on realized limits
  all_preds_long_limits <- all_preds_long_limits %>% 
    mutate(colour = case_when(mean_last_7days > mean_last_7days_min & mean_last_7days < mean_last_7days_max ~ 'inside', T ~ 'outside'))
  
  # Add identifier if predictions are for with or without electrofishing data
  all_preds_long_limits <- all_preds_long_limits %>% 
    mutate(including_electro = T)
  
  # Save output
    saveRDS(all_preds_long_limits %>% 
              mutate(k = set_k), 
            paste0(output_folder, '/temperature_predictions.RDS'))

  
  png(filename = paste0(output_folder, '/thermal_response_curves', '_k=', set_k, '.png'), res = 300, height = 3000, width = 4000)
  print(ggplot(data = all_preds_long_limits %>% 
                 filter(!grepl('Salvelinus_sp_Pro', name))) + 
    geom_line(aes(x = mean_last_7days, y = value, group = name, col = colour)) + 
    scale_colour_manual(values = c('red', 'gray50')) + 
    facet_wrap(~name, scales = 'free_y') + 
    theme_bw() + 
    theme(strip.background = element_blank(), 
          strip.text = element_text(size = 7)))
  dev.off()


  #### MODEL DEPTH PREDICTIONS
  
  ##### GET PREDICTIONS AS STANDARDIZED OBJECT FOR PLOTTING WITH GGPLOT2

model_depth <- expand.grid(
  Depth_sample = seq(
    from = min(df_models_final$Depth_sample, na.rm = TRUE),
    to = max(df_models_final$Depth_sample, na.rm = TRUE), 
    by = 0.02),
  mean_last_7days = mean(df_models_final$mean_last_7days, na.rm = TRUE),
  fProtocol = 'new_protocol', 
  fLake = 'new_lake'
)

# get predictions across model table (model_depth)
preds <- lapply(names(all_gams), function(x) {
  tryCatch({
    pred_x <- predict(
      all_gams[[x]], 
      model_depth, 
      exclude = c('fProtocol', 'fLake'), 
      type = "response", 
      se = TRUE
    )
  }, error = function(e) list(fit = NA, se.fit = NA))
})
  
# process outputs from lapply loop
all_preds <- data.frame(lapply(preds, function(x) x$fit))
names(all_preds) <- names(all_gams)

all_se <- data.frame(lapply(preds, function(x) x$se.fit))
names(all_se) <- names(all_gams)

# bind with predicted values
all_preds_long <- bind_cols(model_depth, all_preds) %>% 
  pivot_longer(5:ncol(.))

all_se_long <- bind_cols(model_depth, all_se) %>% 
  pivot_longer(5:ncol(.), values_to = 'se')

# join together
all_preds_long <- left_join(all_preds_long, all_se_long)

# Calculate the realized limits of the species based on observations
depth_limits <- df_models_final %>% 
  filter(Presence == 1) %>% 
  group_by(Species) %>% 
  summarise(
    across(Depth_sample, list(min = ~min(.x, na.rm = TRUE),
                              max = ~max(.x, na.rm = TRUE)))
  ) %>% 
  rename(name = Species)

# Join in limits
all_preds_long_limits <- left_join(depth_limits, all_preds_long)

# Add logic to colour line depending on realized limits
all_preds_long_limits <- all_preds_long_limits %>% 
  mutate(colour = case_when(
    Depth_sample > Depth_sample_min & Depth_sample < Depth_sample_max ~ 'inside', 
    TRUE ~ 'outside'
  ))

# Add identifier if predictions are for with or without electrofishing data
all_preds_long_limits <- all_preds_long_limits %>% 
  mutate(including_electro = TRUE)

# Save output
saveRDS(all_preds_long_limits %>% 
          mutate(k = set_k), 
        paste0(output_folder, '/depth_predictions.RDS'))

# Plotting
png(filename = paste0(output_folder, '/depth_response_curves', '_k=', set_k, '.png'), res = 300, height = 3000, width = 4000)
print(ggplot(data = all_preds_long_limits %>% 
               filter(!grepl('Salvelinus_sp_Pro', name))) + 
  geom_line(aes(x = Depth_sample, y = value, group = name, col = colour)) + 
  scale_colour_manual(values = c('red', 'gray50')) + 
  facet_wrap(~name, scales = 'free_y') + 
  theme_bw() + 
  theme(strip.background = element_blank(), 
        strip.text = element_text(size = 7)))
dev.off()


  #### SAVE SUMMARIES OF MODEL PROPERTIES
  
  tidy(all_gams[[1]])
  
  tidy_output <- lapply(names(all_gams), function(x){tryCatch({
    
    tidy_x <- tidy(all_gams[[x]])
    tidy_x$dev.exp <- signif(summary(all_gams[[x]])$dev.expl[1],2)
    tidy_x$aic <- all_gams[[x]]$aic
    tidy_x$species_name <- x
    tidy_x
  }, error = function(e) NULL)
    
  })

  table_model_summaries <- bind_rows(tidy_output) %>% 
    relocate(species_name)
  
  write_csv(table_model_summaries, 
            file = paste0(output_folder, '/model_summaries.csv'))
  
  hist(unique(table_model_summaries$dev.exp))
  write_csv(table_model_summaries %>% 
                select(species_name, dev.exp) %>% 
                unique() %>% 
                arrange(dev.exp), file = paste0(output_folder, '/model_devexp.csv')) # really good model deviance explained

  
    hist(unique(table_model_summaries$aic))
  write_csv(table_model_summaries %>% 
                select(species_name, aic) %>% 
                unique() %>% 
                arrange(aic), file = paste0(output_folder, '/model_aic.csv')) # really good model deviance explained

#### END OF LOOPING ACROSS K VALUES

})

```


# Compare response curves across common species 

```{r}

# generate subset of species with abundant data to quantify seperate response curves
many_records <- df_models_final %>% 
  group_by(Species) %>% 
  filter(Presence == 1) %>% 
  tally %>% 
  arrange(desc(n)) 

# read in both response curve sets
preds_no_depth <- readRDS("outputs/run_01052025_electro_k=3/temperature_predictions.RDS") %>% mutate(depth = 'no_depth')
preds_with_depth <- readRDS("outputs/run_01052025_depth_electro_k=3/temperature_predictions.RDS") %>% mutate(depth = 'with_depth')

# combine the depth and no depth models
preds_combined <- bind_rows(
  preds_no_depth, 
  preds_with_depth
) %>% 
  # join and select only species with > 50 records
  left_join(many_records %>% filter(n>=50), 
  ., 
  by = c('Species' = 'name'))

# process names for facets
preds_combined <- 
  preds_combined %>% 
  mutate(Species = gsub('_', ' ', Species))

png(filename = 'figures/thermal_response_curves_DepthComparison_k=3.png', res = 300, height = 3000, width = 3000)
print(ggplot(data = preds_combined %>% 
                 mutate(value = case_when(value > 500 ~ 500, T ~ value))) + 
    geom_line(aes(x = mean_last_7days, y = value, group = depth, col = depth, alpha = log10(n))) + 
    scale_colour_manual(name = 'depth in model', values = c('red', 'blue')) + 
    facet_wrap(~Species, scales = 'free_y', ncol = 5) + 
    theme_bw() + 
    theme(strip.background = element_blank(), 
          strip.text = element_text(size = 7),
          legend.position = 'bottom'))+
    xlab('Temperature') + 
    ylab('Predicted abundance')
dev.off()

```


# Plot which types of models have greater deviance explained 
```{r}

dev_nodepth <- read_csv('outputs/run_01052025_electro_k=3/model_devexp.csv') %>% rename(dev.exp.nodepth = dev.exp)
dev_depth   <- read_csv('outputs/run_01052025_depth_electro_k=3/model_devexp.csv') %>% rename(dev.exp.depth = dev.exp)

dev_exp_data <- left_join(dev_nodepth, dev_depth, by = 'species_name') %>% 
  mutate(delta_dev.exp = dev.exp.depth - dev.exp.nodepth) %>% 
  left_join(., many_records, by = c('species_name' = 'Species'))

png(filename = 'figures/dev_exp_withDepth_k=3.png', res = 300, height = 2000, width = 2000)
ggplot(data = dev_exp_data) + 
  geom_histogram(aes(x = delta_dev.exp*100)) + 
  theme_bw() + 
  geom_vline(aes(xintercept = 0)) + 
  xlab('Gain in deviance explained (%) \n when including depth') + 
  ylab('n. species') + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5))
dev.off()

summary(dev_exp_data)


```


# Read in AIC comparisons
```{r}
aic_nodepth <- read_csv('outputs/run_01052025_electro_k=3/model_aic.csv') %>% rename(aic.nodepth = aic)
aic_depth   <- read_csv('outputs/run_01052025_depth_electro_k=3/model_aic.csv') %>% rename(aic.depth = aic)


aic_data <- left_join(aic_nodepth, aic_depth, by = 'species_name') %>% 
  mutate(delta_aic = aic.depth - aic.nodepth) %>% 
  left_join(., many_records, by = c('species_name' = 'Species'))

aic_data %>% 
  filter(delta_aic < -2)

png(filename = 'figures/aic_withDepth_k=3.png', res = 300, height = 2000, width = 2000)
ggplot(data = aic_data %>% 
         mutate(delta_aic = case_when(delta_aic < -50 ~ -50, T ~ delta_aic))) + 
  geom_histogram(aes(x = delta_aic), bins = 50) + 
  theme_bw() + 
  geom_vline(aes(xintercept = 0)) + 
  geom_vline(aes(xintercept = -2), col = 'red') + 
  xlab('Difference in AIC \n when including depth') + 
  ylab('n. species') + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5))
dev.off()

```


# Plot the ffd from both depth and no-depth models

```{r}

## Get ffd for depth based models
preds_no_depth <- readRDS("outputs/run_01052025_electro_k=3/temperature_predictions.RDS") %>% mutate(depth = 'no_depth')
preds_with_depth <- readRDS("outputs/run_01052025_depth_electro_k=3/temperature_predictions.RDS") %>% mutate(depth = 'with_depth')

# combine the depth and no depth models
preds_combined_v2 <- bind_rows(
  preds_no_depth, 
  preds_with_depth
) %>% 
  left_join(many_records, 
  ., 
  by = c('Species' = 'name')) %>% 
  rename(name = Species)


# get the ffd for all species and depth models
all_ffd_derivs <- preds_combined_v2 %>% 
  group_by(name, k, including_electro, depth) %>% 
  filter(colour == 'inside') %>% 
  mutate(value = scales::rescale(value)) %>% 
  # estimate next values
  mutate(value_lead = lead(value), 
         mean_last_7days_lead = lead(mean_last_7days)) %>% 
  # remove end point
  filter(!is.na(value_lead), !is.na(mean_last_7days_lead)) %>% 
  # forward finite difference
  mutate(ffd = (value_lead - value) / (mean_last_7days_lead - mean_last_7days), 
         ffd_per_0.1degree = (ffd*(1/0.2)) * 0.1)


# pivot to get the ffs per "model depth type"
ffd_comparison <- all_ffd_derivs %>% 
  ungroup() %>% 
  select(name, n, ffd_per_0.1degree, depth, mean_last_7days) %>% 
  pivot_wider(values_from = ffd_per_0.1degree, names_from = depth)
  
ggplot(data = ffd_comparison %>% 
         group_by(name) %>% 
         slice_sample(n=100)) + 
  geom_point(aes(x = no_depth, y = with_depth, col = log10(n))) + 
  facet_wrap(~name, scales = 'free') + 
  scale_color_viridis_c()

# get the rank correlations between different model types
rank_cor <- ffd_comparison %>% 
  arrange(name) %>% 
  group_by(name) %>%
  filter(!is.na(with_depth), !is.na(no_depth)) %>% 
  nest() %>% 
  mutate(rank = purrr::map(data, ~broom::tidy(cor.test(.$no_depth, .$with_depth, method = 'spearman')))) %>% 
  unnest(rank)

# find tendency of rank correlations
rank_cor$estimate %>% summary
hist(rank_cor$estimate)

# Compute summary stats
mean_val <- mean(rank_cor$estimate, na.rm = TRUE)
median_val <- median(rank_cor$estimate, na.rm = TRUE)

# histogram of spearmans rank estimates
png(filename = 'figures/ffd_spearmans_rank.png', res = 300, height = 2000, width = 2000)
ggplot(data = rank_cor, aes(x = estimate)) + 
  geom_histogram(
    bins = 30, 
    fill = "gray50", 
    color = "white"
  ) +
  geom_vline(aes(xintercept = mean_val), color = "black", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = median_val), color = "black", linetype = "solid", size = 1) +
  annotate("text", x = mean_val-0.25, y = 30, label = paste0("Mean=", signif(mean_val,2)), 
           vjust = -0.5, hjust = -0.1, color = "black", size = 3) +
  annotate("text", x = median_val-0.25, y = 30, label = paste0("Median=", signif(median_val,2)), 
           vjust = -0.5, hjust = -0.1, color = "black", size = 3) +
  theme_bw(base_size = 12) +
  xlab("Spearman's rank estimate") +
  ylab("Number of species") +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank()
  )
dev.off()

# plot against number of records
rank_cor <- rank_cor %>% 
  unnest(data) %>% 
  select(name, n, estimate:alternative) %>% 
  unique()


round((table(rank_cor$estimate > 0) / length(rank_cor$estimate)) * 100)
round((table(rank_cor$estimate > 0.9) / length(rank_cor$estimate)) * 100)

# Rescale estimate to [0,1] since quasibinomial expects it
rank_cor_trans <- rank_cor %>%
  mutate(log_n = log10(n), 
         estimate_scaled = (estimate + 1) / 2)

# Step 2: Fit a glm with quasibinomial family
model <- glm(estimate_scaled ~ log_n, data = rank_cor_trans, family = quasibinomial())

# Step 3: Check summary to assess significance
summary(model)

# Plot result
png(filename = 'figures/ffd_depth_nodepth_comparison.png', res = 300, height = 1500, width = 1500)
ggplot(rank_cor_trans, aes(x = log_n, 
                           y = estimate_scaled)) + 
  geom_point() + 
  stat_smooth(method = "glm", method.args = list(family = "quasibinomial"), lty = 2, se = F) +
  scale_y_continuous(
    name = "Correlation estimate",
    breaks = seq(0, 1, 0.1),
    labels = function(x) round(x * 2 - 1, 2)  # Back-transform to [-1, 1]
  ) +
  xlab("log10(n)") +
  theme_bw()
dev.off()


```



```{r}
 
# find those species with negative relationships between thermal ffd when including depth and excluding depth

library(ggh4x)  # for custom strip backgrounds
library(scales) # for rescale()

# Prepare neg_rank as before
neg_rank <- rank_cor %>%
  #filter(estimate < 0) %>%
  mutate(Species = gsub('_', ' ', name)) %>%
  arrange(estimate)

species_estimates <- rank_cor %>%
  mutate(Species = gsub('_', ' ', name)) %>%
  select(Species, estimate)

# Add n (presence records) from many_records to species_estimates (assuming many_records has Species and n)
# Join presence records 'n' to preds_combined and then make label

preds_combined <- bind_rows(preds_no_depth, preds_with_depth) %>%
  rename(Species = name) %>% 
  left_join(many_records, by = "Species") %>%      # make sure many_records has Species and n
  mutate(Species = gsub('_', ' ', Species)) %>%
  left_join(species_estimates, by = "Species") %>%
  filter(Species %in% neg_rank$Species) %>%
  mutate(
    value = pmin(value, 500),
    Species = factor(Species, levels = unique(neg_rank$Species)),
    facet_label = paste0(Species, ", n=", n)     # combine species name and presence count
  )

# Compute x and y ranges for rectangles per facet label
rect_data <- preds_combined %>%
  group_by(facet_label, estimate) %>%
  summarise(
    xmin = min(mean_last_7days),
    xmax = max(mean_last_7days),
    ymin = min(value),
    ymax = max(value),
    .groups = "drop"
  ) %>%
  mutate(
    fill_color = scales::col_numeric(
      palette = c("red", "white", "blue"),
      domain = range(rank_cor$estimate, na.rm = TRUE)
    )(estimate)
  )

png(filename = 'figures/thermal_response_curves_DepthComparison_k=3_RANK.png', res = 300, height = 3000, width = 3000)

print(
  ggplot() +
    geom_rect(data = rect_data,
              aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = fill_color),
              inherit.aes = FALSE, alpha = 0.3) +
    scale_fill_identity() +
    geom_line(data = preds_combined,
              aes(x = mean_last_7days, y = value, group = depth, color = depth, lty = depth)) +
    scale_colour_manual(name = 'Depth in model', values = c('gray30', 'black')) +
    scale_linetype(name = 'Depth in model') + 
    facet_wrap(~facet_label, scales = 'free_y', ncol = 5) +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      strip.text = element_text(size = 7),
      legend.position = 'bottom'
    ) +
    xlab('Temperature') +
    ylab('Predicted abundance')
)

dev.off()

```












