---
title: ""
output: html_document
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

# packages relevant for data analysis
pacman::p_load(
  tidyverse,
  here,
  readr,
  readxl,
  gratia,
  mgcv,
  forcats,
  ggpubr,
  RColorBrewer,
  broom,
  grid,
  DHARMa,
  mgcViz,
  gridExtra,
  viridis, 
  modelsummary,
  gratia, 
  broom,
  tidyverse,
  DHARMa,
  mgcv,
  gratia,
  mgcViz, 
  data.table, 
  ggrepel, 
  cowplot, 
  patchwork
)

```
# Plot response diversity across temperature in each lake
Here we want to visulaise the disimilarity in responses amongst species for each lake

```{r}

all_response_div <- readRDS('data/response_diversity.rds')


dir.create('figures/response_diversity_summaries/')


# dissimilarity
png('figures/response_diversity_summaries/dissimilarity_lakes_facet.png', res = 300, width = 3000, height = 3000)
ggplot(data = all_response_div) + 
  geom_line(aes(x = mean_last_7days, y = dissimilarity_ross, colour = including_electro), alpha = 0.5) + 
  facet_wrap(~lake) + 
  theme_bw() + 
  ylab('Response dissimilarity') + 
  xlab('Temperature') + 
  scale_colour_discrete('with electrofishing')
dev.off()

ggplot(data = all_response_div %>% 
         filter(including_electro == T, k == 3)) + 
  geom_line(aes(x = mean_last_7days, y = dissimilarity_ross, colour = as.character(k))) + 
  facet_wrap(~lake, scales = 'free')


# divergence
png('figures/response_diversity_summaries/divergence_lakes_facet.png', res = 300, width = 3000, height = 3000)
ggplot(data = all_response_div ) + 
  geom_line(aes(x = mean_last_7days, y = divergence_ross, colour = including_electro), alpha = 0.5) + 
  facet_wrap(~lake, scales = 'free') + 
  theme_bw() + 
  ylab('Response divergence') + 
  xlab('Temperature') + 
  scale_colour_discrete('with electrofishing')
dev.off()

ggplot(data = all_response_div %>% 
         filter(including_electro == T)) + 
  geom_line(aes(x = mean_last_7days, y = divergence_ross, colour = as.character(k))) + 
  facet_wrap(~lake, scales = 'free')

ggplot(data = all_response_div %>% 
         filter(including_electro == T, k == 3)) + 
  stat_smooth(aes(x = mean_last_7days, y = divergence_ross, colour = as.character(lake))) 


# sd
png('figures/response_diversity_summaries/sd_ffd_lakes_facet.png', res = 300, width = 3000, height = 3000)
ggplot(data = all_response_div) + 
  geom_line(aes(x = mean_last_7days, y = sd_ffd, colour = including_electro), alpha = 0.5) + 
  facet_wrap(~lake, scales = 'free') + 
  theme_bw() + 
  ylab('Response SD') + 
  xlab('Temperature') + 
  scale_colour_discrete('with electrofishing')
dev.off()

ggplot(data = all_response_div %>% 
         filter(including_electro == T, k == 3)) + 
  geom_line(aes(x = mean_last_7days, y = sd_ffd, colour = as.character(lake))) #+ 
  #facet_wrap(~lake)

ggplot(data = all_response_div %>% 
         filter(including_electro == T, k == 3)) + 
  stat_smooth(aes(x = mean_last_7days, y = sd_ffd, colour = as.character(lake))) #+ 

ggplot(data = all_response_div %>% 
         filter(including_electro == T, k == 3)) + 
  geom_point(aes(x = mean_last_7days, y = sd_ffd, colour = as.character(lake))) #+ 


# mean
png('figures/response_diversity_summaries/mean_ffd_lakes_facet.png', res = 300, width = 3000, height = 3000)
ggplot(data = all_response_div) + 
  geom_line(aes(x = mean_last_7days, y = mean_ffd, colour = including_electro), alpha = 0.5) + 
  facet_wrap(~lake, scales = 'free') + 
  theme_bw() + 
  ylab('Response SD') + 
  xlab('Temperature') + 
  scale_colour_discrete('with electrofishing')
dev.off()

ggplot(data = all_response_div %>% 
         filter(including_electro == T, k == 3) %>% 
         arrange(mean_last_7days)) + 
  geom_line(aes(x = mean_last_7days, y = mean_ffd, colour = as.character(lake))) + 
 # facet_wrap(~lake, scales = 'free') + 
  geom_hline(aes(yintercept = 0)) + 
  scale_y_continuous(limits = c(-0.1, 0.1))

```



```{r}

# Dissimilarity
diss_plot <- ggplot(data = all_response_div %>% 
         filter(k == 3, including_electro == T)) +
  geom_line(aes(x = mean_last_7days, y = dissimilarity_ross, colour = as.character(lake), 
                lty = including_electro), se = F, lwd = 0.5)  + 
  stat_smooth(aes(x = mean_last_7days, y = dissimilarity_ross, 
                lty = including_electro), se = F, col = 'black', lwd = 2) + 
  theme_bw()+ 
  theme(legend.position = "none", 
        aspect.ratio = 1, 
        panel.grid = element_blank()) + 
  xlab('Temperature') + 
  ylab('Response dissimilarity')

# Divergence
div_plot <- ggplot(data = all_response_div %>% 
         filter(k == 3, including_electro == T)) +
  geom_line(aes(x = mean_last_7days, y = divergence_ross, colour = as.character(lake), 
                lty = including_electro), se = F, lwd = 0.5)  + 
  stat_smooth(aes(x = mean_last_7days, y = divergence_ross, 
                lty = including_electro), se = F, col = 'black', lwd = 2) + 
  theme_bw()+ 
  theme(legend.position = "none", 
        aspect.ratio = 1, 
        panel.grid = element_blank()) +
  xlab('Temperature') + 
  ylab('Response divergence')

# SD derivs
sd_plot <- ggplot(data = all_response_div %>% 
         filter(k == 3, including_electro == T)) +
  geom_line(aes(x = mean_last_7days, y = sd_ffd, colour = as.character(lake), 
                lty = including_electro), se = F, lwd = 0.5)  + 
  stat_smooth(aes(x = mean_last_7days, y = sd_ffd, 
                lty = including_electro), se = F, col = 'black', lwd = 2) + 
  theme_bw()+ 
  theme(legend.position = "none", 
        aspect.ratio = 1, 
        panel.grid = element_blank()) + 
    xlab('Temperature') + 
  ylab('Response SD')



# Mean derivs
mean_plot <- ggplot(data = all_response_div %>% 
         filter(k == 3, including_electro == T)) +
  geom_line(aes(x = mean_last_7days, y = mean_ffd, colour = as.character(lake), 
                lty = including_electro), se = F, lwd = 0.5)  + 
  stat_smooth(aes(x = mean_last_7days, y = mean_ffd, 
                lty = including_electro), se = F, col = 'black', lwd = 2) + 
  theme_bw() + 
  geom_hline(aes(yintercept = 0)) + 
  theme(legend.position = "none", 
        aspect.ratio = 1, 
        panel.grid = element_blank()) + 
  xlab('Temperature') + 
  ylab('Response direction')


library(cowplot)

dir.create('figures/response_diversity_summaries/')
png('figures/response_diversity_summaries/all_responses.png', res = 300, width = 3000, height = 1500)
cowplot::plot_grid(diss_plot, sd_plot, div_plot, mean_plot, ncol = 1)
dev.off()

```

# Lake metadata and species occurrence information

```{r}

# lake metadata
lake_meta <- read_xlsx('data/lake_info.xlsx') %>% 
  select(Lake, Drainage, Altitude, Lake_area, Lake_volume, Max_depth, contains('Phos_')) %>% 
  janitor::clean_names() %>% 
  mutate(across(contains('phos_'), as.numeric)) %>% 
  mutate(lake = case_when(lake %in% c('ConstanceObersee', 'ConstanceUntersee') ~ 'Constance',
                          lake %in% c('ZurichObersee', 'ZurichUntersee') ~ 'Zurich',
                          T ~ lake))

# summarise where no information in subbasin
lake_meta <- lake_meta %>% 
  group_by(lake) %>% 
  summarise(across(is.numeric, max, na.rm = T)) 


# information about categorization in each lake for the subgroups per lake
# get per lake biogeographic affinity
biogeography_swisswide <- read_csv("data/taxonomy_May2025/species_lake_list_raw_EDITED.csv") %>% 
  select(Species_ecotype, Lake, Swisswide_biogeographic_status, Lake_biogeographic_status) %>% 
  rename(lake = Lake, sbs = Swisswide_biogeographic_status, lbs = Lake_biogeographic_status) %>% 
  unique() %>% 
  filter(lbs != 'remove') %>% 
  mutate(lbs_intro = case_when(sbs == 'native' & lbs == 'non_native' ~ 'native_intro', 
                               sbs == 'endemic' & lbs == 'non_native' ~ 'endemic_intro', 
                               T ~ lbs))

# biogeography_swisswide %>% print(n = 500)

# lake richness
lake_richness <- biogeography_swisswide %>% 
  group_by(lake) %>% 
  summarise(lake_richness = n_distinct(Species_ecotype))

# group richness
group_richness <- biogeography_swisswide %>% 
  group_by(lake, lbs_intro) %>% 
  summarise(lake_richness = n_distinct(Species_ecotype)) %>% 
  pivot_wider(values_from = lake_richness, names_from = lbs_intro, values_fill = 0) %>% 
  ungroup()
  


# join lake and group richness to response diversity data
result <- reduce(list(lake_meta, lake_richness, group_richness), ~ left_join(.x, .y, by = "lake"))


```

# Plots of average response diversity across whole lake

```{r}

# match response diversity information with lake
final_rd <- all_response_div %>% 
  filter(k == 3, including_electro == T)


# Aggregate the 'final_rd' data frame for each response diversity metric by 'lake'
rd_aggregate <- final_rd %>% 
  # Group the data by the 'lake' column
  group_by(lake) %>% 
  # For all columns from 'dissimilarity_ross' to 'sd_ffd', compute the mean
  # 'na.rm = T' ensures that NA values are ignored during the mean calculation
  summarise(across(dissimilarity_ross:sd_ffd, mean, na.rm = T)) %>% 
  # Join the summarised data with the 'result' data frame using a left join
  left_join(result)


# generalized function to plot variables by choice
plot_lake_relationship <- function(data, xvar, yvar) {

  # Compute Spearman correlation
  COR <- cor.test(y = data[[yvar]], x = data[[xvar]], method = 'spearman')

  # Extract and format stats
  rho <- round(as.numeric(COR$estimate), 2)
  pval <- signif(COR$p.value, 2)
  
  # Standard significance label
  p_sig <- case_when(
    COR$p.value <= 0.001 ~ "p < 0.001",
    COR$p.value <= 0.01  ~ "p < 0.01",
    COR$p.value <= 0.05  ~ "p < 0.05",
    TRUE          ~ "p > 0.05"
  )
  
  # Calculate axis ranges
  x_max <- min(data[[xvar]], na.rm = TRUE)
  y_max <- max(data[[yvar]], na.rm = TRUE)

  ggplot(data = data) + 
    stat_smooth(aes(x = .data[[xvar]], y = .data[[yvar]]), 
                method = 'lm', alpha = 0.5, col = 'gray20', fill = 'gray80', 
                lty = case_when(COR$p.value < 0.05 ~ 1, T ~ 2),
                se  = case_when(COR$p.value < 0.05 ~ F, T ~ F)) + 
    geom_text_repel(aes(x = .data[[xvar]], y = .data[[yvar]], label = lake, col = lake), max.overlaps = 5) + 
    geom_point(aes(x = .data[[xvar]], y = .data[[yvar]], col = lake)) + 
    theme_bw() + 
    theme(
      aspect.ratio = 1, 
      panel.grid = element_blank(),
      legend.position = 'none',
      plot.title = element_text(size = 12) 
    ) + 
    ggtitle(paste0("ρ = ", rho, ", ", p_sig))
}


# plots for lake richness
rich_diss    <- plot_lake_relationship(rd_aggregate, 'lake_richness', 'dissimilarity_ross') + ylab('Response dissimilarity') + xlab('Species richness') 
rich_div     <- plot_lake_relationship(rd_aggregate, 'lake_richness', 'divergence_ross') + ylab('Response divergence') + xlab('Species richness') 
rich_sdffd   <- plot_lake_relationship(rd_aggregate, 'lake_richness', 'sd_ffd') + ylab('Response SD') + xlab('Species richness') 
rich_meanffd <- plot_lake_relationship(rd_aggregate, 'lake_richness', 'mean_ffd') + ylab('Response direction') + xlab('Species richness') 
png('figures/response_diversity_summaries/richness_wholeLake.png', res = 300, width = 4000, height = 3000)
print(plot_grid(rich_diss, 
                rich_div, 
                rich_sdffd, 
                rich_meanffd, nrow = 2, ncol = 2))
dev.off()

```

# combine together aggregated richness with across-gradient plot
```{r}

png('figures/response_diversity_summaries/temp_richness_wholeLake.png', res = 600, width = 7000, height = 4000)
plot_grid(diss_plot ,
          sd_plot , 
          div_plot ,
          mean_plot , 
          ncol = 4) / 
  plot_grid(rich_diss, 
            rich_sdffd,
            rich_div, 
            rich_meanffd ,
            ncol = 4) + 
plot_layout(guides = "collect")
dev.off()


```


# Maybe moving windows of 2 degrees are more intuitive

```{r}

all_response_div_round <- all_response_div %>% 
  mutate(mean_last_7days = round(mean_last_7days)) %>% 
  group_by(mean_last_7days, k, including_electro, lake) %>% 
  mutate(across(dissimilarity_ross:sd_ffd, mean)) %>% 
  unique()

```

# Plot running mean metrics for easier visualisation

```{r}


# dissimilarity
ggplot(data = all_response_div_round %>% 
         filter(k == 3)) + 
  stat_smooth(aes(x = mean_last_7days, y = dissimilarity_ross, 
                  colour = as.character(including_electro))) + 
  facet_wrap(~lake)

ggplot(data = all_response_div_round %>% 
         filter(including_electro == T, k == 3)) + 
    geom_point(aes(x = mean_last_7days, y = dissimilarity_ross)) +
  stat_smooth(aes(x = mean_last_7days, y = dissimilarity_ross, colour = as.character(lake)), se = F) 


# divergence
ggplot(data = all_response_div_round %>% 
         filter(k == 3)) + 
  stat_smooth(aes(x = mean_last_7days, y = divergence_ross, colour = including_electro)) + 
  facet_wrap(~lake, scales = 'free')

ggplot(data = all_response_div_round %>% 
         filter(including_electro == T, k != 10)) + 
  stat_smooth(aes(x = mean_last_7days, y = divergence_ross, colour = as.character(k))) + 
  facet_wrap(~lake, scales = 'free')


# sd
ggplot(data = all_response_div_round %>% 
         filter(k == 3)) + 
  stat_smooth(aes(x = mean_last_7days, y = sd_ffd, colour = including_electro)) + 
  facet_wrap(~lake)

# mean
ggplot(data = all_response_div_round %>% 
         filter(k == 3)) + 
  geom_line(aes(x = mean_last_7days, y = mean_ffd, colour = including_electro)) + 
  facet_wrap(~lake, scales = 'free')

ggplot(data = all_response_div_round %>% 
         filter(!k%in%c(2, 10))) +
  geom_line(aes(x = mean_last_7days, y = mean_ffd, colour = as.character(k), 
                lty = including_electro)) + 
  facet_wrap(~lake, scales = 'free') + 
  geom_hline(aes(yintercept = 0))


```



