---
title: "04_model_RD"
output:
  html_document:
    css: styles.css
date: "`r Sys.Date()`"
author: "Sophie Moreau, Conor Waldock"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, echo = F}
knitr::opts_chunk$set(echo = TRUE)

# packages relevant for data analysis
pacman::p_load(
  tidyverse,
  here,
  readr,
  readxl,
  gratia,
  mgcv,
  forcats,
  ggpubr,
  RColorBrewer,
  broom,
  grid,
  DHARMa,
  mgcViz,
  gridExtra,
  viridis, 
  modelsummary,
  gratia, 
  broom,
  tidyverse,
  DHARMa,
  mgcv,
  gratia,
  mgcViz, 
  ggrepel, 
  rlang, 
  cowplot, 
  patchwork
)

```

# Fit models explaining variation in response diversity based on lake properties

We aim to answer:

1.  Is there a association between species richness per lake on response diversity?

2.  Is there an association between species richness of different biogeographic groups on response diversity, does the strength of this association differ between groups?

3.  Is there an effect of lake eutrophical status on response diversity?

```{r}

# read in response diversity data
all_response_div <- readRDS('data/response_diversity.rds')

# match response diversity information with lake
final_rd <- all_response_div %>% 
  filter(k == 3, including_electro == T)

# lake metadata
lake_meta <- read_xlsx('data/lake_info.xlsx') %>% 
  select(Lake, Drainage, Altitude, Lake_area, Lake_volume, Max_depth, contains('Phos_')) %>% 
  janitor::clean_names() %>% 
  mutate(across(contains('phos_'), as.numeric)) %>% 
  mutate(lake = case_when(lake %in% c('ConstanceObersee', 'ConstanceUntersee') ~ 'Constance',
                          lake %in% c('ZurichObersee', 'ZurichUntersee') ~ 'Zurich',
                          T ~ lake))

# summarise where no information in subbasin
lake_meta <- lake_meta %>% 
  group_by(lake) %>% 
  summarise(across(is.numeric, max, na.rm = T)) 


# information about categorization in each lake for the subgroups per lake
# get per lake biogeographic affinity
biogeography_swisswide <- read_csv("data/taxonomy_May2025/species_lake_list_raw_EDITED.csv") %>% 
  select(Species_ecotype, Lake, Swisswide_biogeographic_status, Lake_biogeographic_status) %>% 
  rename(lake = Lake, sbs = Swisswide_biogeographic_status, lbs = Lake_biogeographic_status) %>% 
  unique() %>% 
  filter(lbs != 'remove') %>% 
  mutate(lbs_intro = case_when(sbs == 'native' & lbs == 'non_native' ~ 'native_intro', 
                               sbs == 'endemic' & lbs == 'non_native' ~ 'endemic_intro', 
                               T ~ lbs))

# biogeography_swisswide %>% print(n = 500)

# lake richness
lake_richness <- biogeography_swisswide %>% 
  group_by(lake) %>% 
  summarise(lake_richness = n_distinct(Species_ecotype))

# group richness
group_richness <- biogeography_swisswide %>% 
  group_by(lake, lbs_intro) %>% 
  summarise(lake_richness = n_distinct(Species_ecotype)) %>% 
  pivot_wider(values_from = lake_richness, names_from = lbs_intro, values_fill = 0) %>% 
  ungroup()
  
# join lake and group richness to response diversity data
result <- reduce(list(lake_meta, lake_richness, group_richness), ~ left_join(.x, .y, by = "lake"))

```

# Average response diversity per lake

Here I aggregate the response diversity across the whole temperature profile of the lake. This mixes the response diversity of the deeper waters and shallower waters in ways that may be difficult to interpret. Later we will not aggregate the data in this way but instead seperate out the include across all lakes.

```{r}

# Aggregate the 'final_rd' data frame for each response diversity metric by 'lake'
rd_aggregate <- final_rd %>% 
  # Group the data by the 'lake' column
  group_by(lake) %>% 
  # For all columns from 'dissimilarity_ross' to 'sd_ffd', compute the mean
  # 'na.rm = T' ensures that NA values are ignored during the mean calculation
  summarise(across(dissimilarity_ross:sd_ffd, mean, na.rm = T)) %>% 
  # Join the summarised data with the 'result' data frame using a left join
  left_join(result)


# generalized function to plot variables by choice
plot_lake_relationship <- function(data, xvar, yvar) {
  ggplot(data = data) + 
    stat_smooth(aes(x = {{ xvar }}, y = {{ yvar }}), 
                method = 'lm', alpha = 0.5, col = 'gray20', fill = 'gray80') + 
    geom_text_repel(aes(x = {{ xvar }}, y = {{ yvar }}, label = lake)) + 
    geom_point(aes(x = {{ xvar }}, y = {{ yvar }}, col = lake)) + 
    theme_bw() + 
    theme(
      aspect.ratio = 1, 
      panel.grid = element_blank(),
      legend.position = 'none'
    )
}

# plot but split across endemic groupings
plot_lake_relationship_endemic <- function(data, xvar, yvar, grouping) {
  ggplot(data = data) + 
    stat_smooth(aes(x = {{ xvar }}, y = {{ yvar }}, group = {{ grouping }}, colour = {{grouping}}), 
                method = 'lm', alpha = 0.5, fill = 'gray80') + 
    # geom_text_repel(aes(x = {{ xvar }}, y = {{ yvar }}, label = lake)) + 
    geom_point(aes(x = {{ xvar }}, y = {{ yvar }}, colour = {{grouping}})) + 
    theme_bw() + 
    theme(
      aspect.ratio = 1, 
      panel.grid = element_blank()
    )
}

# plots for lake richness
rich_diss    <- plot_lake_relationship(rd_aggregate, lake_richness, dissimilarity_ross) + ylab('Response Dissimilarity') + xlab('Species richness') 
rich_div     <- plot_lake_relationship(rd_aggregate, lake_richness, divergence_ross) + ylab('Response Divergence') + xlab('Species richness') 
rich_sdffd   <- plot_lake_relationship(rd_aggregate, lake_richness, sd_ffd) + ylab('Response SD') + xlab('Species richness') 
rich_meanffd <- plot_lake_relationship(rd_aggregate, lake_richness, mean_ffd) + ylab('Response Mean') + xlab('Species richness') 
png('figures/response_diversity_summaries/richness_wholeLake.png', res = 300, width = 4000, height = 3000)
print(plot_grid(rich_diss, rich_div, rich_sdffd, rich_meanffd, nrow = 2, ncol = 2))
dev.off()

# plots for endemic richness
rd_aggregate_class <- rd_aggregate %>% pivot_longer(cols = endemic:non_native, names_to = 'biogeographic_class', values_to = 'richness_bioclass')
endm_diss    <- plot_lake_relationship_endemic(rd_aggregate_class, richness_bioclass, dissimilarity_ross, biogeographic_class)
endm_div     <- plot_lake_relationship_endemic(rd_aggregate_class, richness_bioclass, divergence_ross, biogeographic_class)
endm_sdffd   <- plot_lake_relationship_endemic(rd_aggregate_class, richness_bioclass, sd_ffd, biogeographic_class)
endm_meanffd <- plot_lake_relationship_endemic(rd_aggregate_class, richness_bioclass, mean_ffd, biogeographic_class)
png('figures/response_diversity_summaries/endemic_richness_wholeLake.png', res = 300, width = 4000, height = 3000)
plot_grid(endm_diss, endm_div, endm_sdffd, endm_meanffd, nrow = 2, ncol = 2)
dev.off()


rd_aggregate_class <- rd_aggregate %>% pivot_longer(cols = endemic:non_native, names_to = 'biogeographic_class', values_to = 'richness_bioclass')
endm_diss    <- plot_lake_relationship_endemic(rd_aggregate_class %>% filter(biogeographic_class %in% c('endemic', 'native')), richness_bioclass, dissimilarity_ross, biogeographic_class)
endm_div     <- plot_lake_relationship_endemic(rd_aggregate_class %>% filter(biogeographic_class %in% c('endemic', 'native')), richness_bioclass, divergence_ross, biogeographic_class)
endm_sdffd   <- plot_lake_relationship_endemic(rd_aggregate_class %>% filter(biogeographic_class %in% c('endemic', 'native')), richness_bioclass, sd_ffd, biogeographic_class)
endm_meanffd <- plot_lake_relationship_endemic(rd_aggregate_class %>% filter(biogeographic_class %in% c('endemic', 'native')), richness_bioclass, mean_ffd, biogeographic_class)
png('figures/response_diversity_summaries/endemic_richness_v2_wholeLake.png', res = 300, width = 4000, height = 3000)
plot_grid(endm_diss, endm_div, endm_sdffd, endm_meanffd, nrow = 2, ncol = 2)
dev.off()


```

### Using spearmans rank correlation to test simple relationships between variables

In this analysis we have very few degrees of freedom and only 16 lakes. Fitting models with interaction terms is not likely to be revealing. Instead we can see if the correlations are similar amongst different groups and compare the group richness this way. This is likely more conservative to avoid falsely rejecting hypothesis that may be difficult to test with more sophisticated interaction terms.

```{r}

# Define independent (x) and dependent (y) variables
independent_vars <- c("dissimilarity_ross", "divergence_ross", "sd_ffd", "mean_ffd")  # Add more as needed
dependent_vars <- c("lake_richness", "endemic_intro", "endemic", "native", "native_intro", "non_native", "phos_max")

# Initialize an empty list to store results
all_cor_results <- list()

# Nested loop: outer for independent vars, inner for dependent vars
for (x_var in independent_vars) {
  for (y_var in dependent_vars) {
    # Perform correlation test
    test <- cor.test(rd_aggregate[[x_var]], rd_aggregate[[y_var]], method = 'spearman')
    
    # Store results
    all_cor_results[[paste(x_var, y_var, sep = "_vs_")]] <- data.frame(
      x_var = x_var,
      y_var = y_var,
      spearmans_rank = round(test$estimate, 2),
      p_value = round(test$p.value,3)
    )
  }
}

# Combine all results into a single summary data frame
cor_summary_table <- do.call(rbind, all_cor_results)
rownames(cor_summary_table) <- NULL

# View the summary table
print(cor_summary_table)

write_csv(cor_summary_table, 'figures/response_diversity_Figs/cor_biog_temp.csv')

```

## Assessing response diversity across thermal gradients

Next we wish to aggregate response diversity metrics at different temperature bands across the lakes and test for the strength of these relationships. Here we will plot the response diversity with richness relationships across each lake for each temperature bin. I have binned the tempeatures to 5 degree windows (defined by the nearest 5 degree interval (e.g., 2.5 - 7.5 = 5, 7.5 - 12.5 = 10 etc.) and mean-averaged the response diversity values.

```{r}

# Here we want to aggregate by temperature bins
rd_aggregate_tb <- final_rd %>% 
  # Create temperature bins 
  mutate(temp_bin = paste0('temp_', sprintf("%02d", plyr::round_any(mean_last_7days, accuracy = 5)))) %>% 
  # Group the data by the 'lake' column
  group_by(lake, temp_bin) %>% 
  # For all columns from 'dissimilarity_ross' to 'sd_ffd', compute the mean
  # 'na.rm = T' ensures that NA values are ignored during the mean calculation
  summarise(across(dissimilarity_ross:sd_ffd, mean, na.rm = T)) %>% 
  # Join the summarised data with the 'result' data frame using a left join
  left_join(result) %>% 
  # Temp 25 only has zug and maggiore so remove
  filter(temp_bin != 'temp_25')


# Define the variables to use instead of `lake_richness`
richness_vars <- c("lake_richness", "endemic", "native", "endemic_intro", "native_intro", "non_native")  
dissimilarity_vars <- c("dissimilarity_ross", "divergence_ross", "sd_ffd", "mean_ffd")     


# update function to remove text
plot_lake_relationship <- function(data, xvar, yvar) {
  ggplot(data = data) + 
    stat_smooth(aes(x = {{ xvar }}, y = {{ yvar }}), 
                method = 'lm', alpha = 0.5, col = 'gray20', fill = 'gray80') + 
    #geom_text_repel(aes(x = {{ xvar }}, y = {{ yvar }}, label = lake)) + 
    geom_point(aes(x = {{ xvar }}, y = {{ yvar }}, col = lake)) + 
    theme_bw() + 
    theme(
      aspect.ratio = 1, 
      panel.grid = element_blank(),
      legend.position = 'none'
    )
}


# Generate all plots
# Generate all plots
# Generate all plots
all_plots <- map(dissimilarity_vars, function(diss_var) {
  diss_sym <- sym(diss_var)  # convert y-axis var to symbol

  # Loop over each richness (x-axis) variable
  map(richness_vars, function(rich_var) {
    rich_sym <- sym(rich_var)  # convert x-axis var to symbol

    # Loop over temp bins to create individual plots
    plots <- map(unique(rd_aggregate_tb$temp_bin), function(temp_bin_val) {
      data_filtered <- rd_aggregate_tb %>% filter(temp_bin == temp_bin_val)

      plot_lake_relationship(data_filtered, !!rich_sym, !!diss_sym) +
        ggtitle(temp_bin_val) + 
        theme(text = element_text(size = 5))
    })

    # Combine plots for one richness/dissimilarity combo
    wrap_plots(plots, nrow = 1) + plot_annotation(title = diss_sym)

  })
}) %>% flatten()  # flatten from nested lists into a single list

print(all_plots)
```


## Convert the observed results into a publication quality figure 
Lets have a big multiplanel figure for fun

```{r}
rd_corvalues <- rd_aggregate_tb %>% 
  dplyr::filter(lake != 'Poschiavo') %>% 
  pivot_longer(cols = c(phos_max, lake_richness:non_native), names_to = "richness_type", values_to = "richness_value") %>% 
  pivot_longer(cols = dissimilarity_ross:sd_ffd, names_to = "dissimilarity_type", values_to = "diss_value") %>% 
  group_by(temp_bin, richness_type, dissimilarity_type) %>% 
  summarise(
    cor_result = list(cor.test(diss_value, richness_value)),
    .groups = "drop"
  ) %>%
  mutate(
    estimate = map_dbl(cor_result, ~ round(.x$estimate, 2)),
    p_value = map_dbl(cor_result, ~ signif(.x$p.value, 3))
  ) 

# create plot data type
plot_data_RD <- rd_aggregate_tb %>% 
  dplyr::filter(lake != 'Poschiavo') %>% 
  pivot_longer(cols = c(lake_richness:non_native), names_to = "richness_type", values_to = "richness_value") %>% 
  pivot_longer(cols = dissimilarity_ross:sd_ffd, names_to = "dissimilarity_type", values_to = "diss_value") %>% 
  left_join(rd_corvalues)


plot_data_RD <- plot_data_RD %>% 
  ungroup() %>% 
  mutate(richness_type = dplyr::recode(richness_type, 
                                       'lake_richness' = 'all species',
                                       'native' = 'widespread \n native',
                                       'endemic' = 'in situ \n diversification (ISD)',
                                       'non_native' = 'non-native',
                                       'native_intro' = 'introduced \n native',
                                       'endemic_intro' = 'introduced ISD'), 
         temp_bin  = dplyr::recode(temp_bin, 
                                   'temp_05' = '2.5° to 7.5°',
                                   'temp_10' = '7.5° to 12.5°',
                                   'temp_15' = '12.5° to 17.5°',
                                   'temp_20' = '17.5° to 22.5°'))

# relevel factors
sort(unique(plot_data_RD$richness_type))
plot_data_RD$richness_type <- factor(
  plot_data_RD$richness_type,
  levels = sort(unique(plot_data_RD$richness_type))[c(1,6,2,5,3,4)]
)

plot_data_RD$temp_bin <- factor(
  plot_data_RD$temp_bin,
  levels = sort(unique(plot_data_RD$temp_bin))[rev(c(3, 4, 1, 2))]
)

# rescale to per richness gain - this is interesting but makes it more unintuitive to understand
# plot_data_RD <- plot_data_RD %>% 
#   group_by(temp_bin, richness_type, dissimilarity_type) %>% 
#   mutate(richness_value = richness_value - min(richness_value))


# Confirm levels
levels(plot_data_RD$richness_type)

my_custom_theme <- function() {
  theme_bw() +
    theme(
      panel.grid = element_blank(),
      legend.position = 'none',
      strip.background = element_blank(), 
      aspect.ratio = 1,
      strip.text = element_text(size = 7), 
      text = element_text(size = 8)
    )
}


# Then plot (no change needed here)
ggplot(data = plot_data_RD %>%
         filter(dissimilarity_type == 'dissimilarity_ross')) +
  geom_point(aes(x = richness_value, y = diss_value)) +
  stat_smooth(
    aes(
      x = richness_value, y = diss_value,
      col = richness_type,
      lty = case_when(p_value < 0.05 ~ '1', TRUE ~ '2'),
      fill = case_when(p_value < 0.05 ~ 'gray50', TRUE ~ 'transparent')
    ),
    method = 'lm'
  ) +
  facet_grid(temp_bin ~ richness_type) + 
  my_custom_theme() + 
  scale_fill_manual(values = c('gray50', 'transparent')) +
  ylab('Response Dissimilarity') +
  xlab('Species Richness')


ggplot(data = plot_data_RD %>% 
  filter(dissimilarity_type == 'divergence_ross')) + 
    geom_point(aes(x = richness_value, y = diss_value)) + 
  stat_smooth(aes(x = richness_value, y = diss_value, col = richness_type, 
                  lty = case_when(p_value < 0.05 ~ '1',  T ~ '2'), 
                  fill = case_when(p_value < 0.05 ~ 'gray50',  T ~ 'transparent')), method = 'lm') + 
  facet_grid(temp_bin ~ richness_type) + 
  my_custom_theme() + 
  scale_fill_manual(values = c('gray50', 'transparent')) + 
  ylab('Response Divergence') + 
  xlab('Species Richness')


ggplot(data = plot_data_RD %>% 
         filter(dissimilarity_type == 'sd_ffd')) + 
    geom_point(aes(x = richness_value, y = diss_value)) + 
  stat_smooth(aes(x = richness_value, y = diss_value, col = richness_type, 
                  lty = case_when(p_value < 0.05 ~ '1',  T ~ '2'), 
                  fill = case_when(p_value < 0.05 ~ 'gray50',  T ~ 'transparent')), method = 'lm') + 
  facet_grid(temp_bin ~ richness_type) + 
  my_custom_theme() + 
  scale_fill_manual(values = c('gray50', 'transparent')) + 
  ylab('Response SD') + 
  xlab('Species Richness')


ggplot(data = plot_data_RD %>%
  filter(dissimilarity_type == 'mean_ffd')) + 
    geom_point(aes(x = richness_value, y = diss_value)) + 
  stat_smooth(aes(x = richness_value, y = diss_value, col = richness_type, 
                  lty = case_when(p_value < 0.05 ~ '1',  T ~ '2'), 
                  fill = case_when(p_value < 0.05 ~ 'gray50',  T ~ 'transparent')), method = 'lm') + 
  facet_grid(temp_bin ~ richness_type) + 
  my_custom_theme() + 
  scale_fill_manual(values = c('gray50', 'transparent')) + 
  ylab('Response Mean') + 
  xlab('Species Richness')
```


```{r}
# create plot data type for phosphorous
plot_data_phos <- rd_aggregate_tb %>% 
  dplyr::filter(lake != 'Poschiavo') %>% 
  pivot_longer(cols = c(phos_max), names_to = "richness_type", values_to = "richness_value") %>% 
  pivot_longer(cols = dissimilarity_ross:sd_ffd, names_to = "dissimilarity_type", values_to = "diss_value")%>% 
         left_join(rd_corvalues)

rd_corvalues %>% 
  filter(richness_type == 'phos_max')


plot_data_phos <- plot_data_phos %>% 
  ungroup() %>% 
  mutate(dissimilarity_type = dplyr::recode(dissimilarity_type, 
                                       'dissimilarity_ross' = 'response dissimilarity',
                                       'divergence_ross' = 'response divergence',
                                       'sd_ffd' = 'response SD',
                                       'mean_ffd' = 'response direction'), 
         temp_bin  = dplyr::recode(temp_bin, 
                                   'temp_05' = '2.5° to 7.5°',
                                   'temp_10' = '7.5° to 12.5°',
                                   'temp_15' = '12.5° to 17.5°',
                                   'temp_20' = '17.5° to 22.5°'))

plot_data_phos$dissimilarity_type <- factor(
  plot_data_phos$dissimilarity_type,
  levels = sort(unique(plot_data_phos$dissimilarity_type))[c(2,4,3,1)]
)


unique(plot_data_phos$temp_bin)
plot_data_phos$temp_bin <- factor(
  plot_data_phos$temp_bin,
  levels = sort(unique(plot_data_phos$temp_bin))[c(3,4,1,2)]
)

# Create a label for each facet
facet_labels <- plot_data_phos %>%
  group_by(dissimilarity_type, temp_bin) %>%
  summarise(
    estimate = unique(round(estimate, 2)),
    p_value = unique(p_value),
    .groups = "drop"
  ) %>%
  mutate(
    p_label = case_when(
      p_value < 0.001 ~ "'***'",
      p_value < 0.01 ~ "'**'",
      p_value < 0.05 ~ "'*'",
      TRUE ~ "' '"
    ),
    label = glue::glue("rho == {estimate}^{{{p_label}}}")
  )


facet_positions <- plot_data_phos %>%
  group_by(dissimilarity_type) %>%
  summarise(
    x = max(richness_value, na.rm = TRUE),
    y = max(diss_value, na.rm = TRUE),
    .groups = "drop"
  )

facet_labels <- left_join(facet_labels, facet_positions,
                          by = c("dissimilarity_type"))

phos_plot <- ggplot(data = plot_data_phos) + 
  geom_point(aes(x = richness_value, y = diss_value)) + 
  # Smooths where p < 0.05, show SE
  stat_smooth(
    data = plot_data_phos %>% filter(p_value < 0.05),
    aes(x = richness_value, y = diss_value, 
        lty = 1, fill = 'gray50', col = estimate),
    method = 'lm',
    lty = 1,
    se = TRUE
  ) +

  # Smooths where p >= 0.05, hide SE
  stat_smooth(
    data = plot_data_phos %>% filter(p_value >= 0.05),
    aes(x = richness_value, y = diss_value, 
        col = estimate),
    method = 'lm',
    lty = 2,
    se = FALSE
  ) +
  geom_text(data = facet_labels,
            aes(x = x, y = y, label = label),
            hjust = 1, vjust = 1, size = 3, fontface = "italic", 
                      parse = TRUE) + 
  facet_grid(dissimilarity_type ~ temp_bin, scales = 'free') + 
  my_custom_theme() + 
  scale_fill_manual(values = c('gray50', 'transparent')) + 
  scale_colour_gradient2() +
  ylab('Response metric') + 
  xlab('Maximum phosphorous')

phos_plot
```


# Plot summarisings all correlations between richness and response diversity

```{r}

rd_corvalues_v2 <- rd_corvalues %>% 
  ungroup() %>% 
  filter(richness_type != 'phos_max') %>% 
  mutate(dissimilarity_type = dplyr::recode(dissimilarity_type, 
                                       'dissimilarity_ross' = 'response dissimilarity',
                                       'divergence_ross' = 'response divergence',
                                       'sd_ffd' = 'response SD',
                                       'mean_ffd' = 'response direction'), 
         richness_type = dplyr::recode(richness_type, 
                                       'lake_richness' = 'all species',
                                       'native' = 'widespread native',
                                       'endemic' = 'in situ diversification (ISD)',
                                       'non_native' = 'non-native',
                                       'native_intro' = 'introduced native',
                                       'endemic_intro' = 'introduced ISD'), 
         temp_bin  = dplyr::recode(temp_bin, 
                                   'temp_05' = '2.5 to 7.5',
                                   'temp_10' = '7.5 to 12.5',
                                   'temp_15' = '12.5 to 17.5',
                                   'temp_20' = '17.5 to 22.5'))

# relevel factors
sort(unique(rd_corvalues_v2$richness_type))
rd_corvalues_v2$richness_type <- factor(
  rd_corvalues_v2$richness_type,
  levels = sort(unique(rd_corvalues_v2$richness_type))[rev(c(1,6,2,5,3,4))]
)

rd_corvalues_v2$temp_bin <- factor(
  rd_corvalues_v2$temp_bin,
  levels = sort(unique(rd_corvalues_v2$temp_bin))[c(3, 4, 1, 2)]
)

rd_corvalues_v2$dissimilarity_type <- factor(
  rd_corvalues_v2$dissimilarity_type,
  levels = sort(unique(rd_corvalues_v2$dissimilarity_type))[c(2,4,3,1)]
)


overall_plot <- ggplot(data = rd_corvalues_v2) + 
  geom_tile(aes(y = richness_type, x = temp_bin, fill = estimate)) + 
  geom_text(aes(y = richness_type, x = temp_bin, label = case_when(p_value > 0.05 ~ '', 
                                                                   p_value < 0.05 & p_value > 0.01 ~ '*', 
                                                                   p_value < 0.01 & p_value > 0.001~ '**', 
                                                                   p_value < 0.001 ~ '***'))) + 
  facet_wrap(~dissimilarity_type) + 
  scale_fill_gradient2('Spearman rank') + 
  xlab('\n Habitat temperature bin') + 
  ylab('Richnes grouping \n') + 
  theme_minimal() + 
  theme(panel.grid = element_blank(), 
        legend.position = 'bottom')


```


# Save final figures manually
```{r, include=FALSE}
dir.create('figures/response_diversity_Figs', recursive = T)

my_custom_theme_V2 <- function() {
  theme_bw() +
    theme(
      panel.grid = element_blank(),
      legend.position = 'none',
      strip.background = element_blank(), 
      aspect.ratio = 1,
      strip.text = element_text(size = 10), 
      text = element_text(size = 10)
    )
}

### Dissimilarity effect
png('figures/response_diversity_Figs/Richness_effect_dissimilarity.png', res = 600, width = 4000, height = 4000)
ggplot(data = plot_data_RD %>%
         filter(dissimilarity_type == 'dissimilarity_ross', 
                richness_type %in% levels(richness_type)[c(1,2,3,4)], 
                temp_bin %in% unique(temp_bin)[c(1,4)]) %>% 
         mutate(fill_condition = case_when(p_value < 0.05 & richness_type == 'widespread \n native' ~ '#C7C35E',
                                           p_value < 0.05 & richness_type == 'in situ \n diversification (ISD)' ~ '#3CB371',
                                           p_value < 0.05 & richness_type == 'non-native' ~ 'purple',
                                           TRUE ~ 'white'))) +
  stat_smooth(
    aes(
      x = richness_value, y = diss_value,
      col = richness_type,
      lty = case_when(p_value < 0.05 ~ '1', TRUE ~ '2'),
      fill = fill_condition,
      group = paste0(richness_type, temp_bin)
    ),
    method = 'lm'
  ) +
  geom_point(aes(x = richness_value, y = diss_value, col = richness_type)) +
  facet_grid(temp_bin ~ richness_type, scales = 'free') + 
  my_custom_theme_V2() + 
  #scale_fill_manual(values = c('gray50', 'transparent')) +
  ylab('Response Dissimilarity') +
  xlab('Species Richness') + 
  scale_fill_manual(values = c("#3CB371",  "#C7C35E", "transparent", 'purple')) + 
  scale_colour_manual(values = c("gray40",  "#C7C35E", "#3CB371", 'purple'))
dev.off()

# Plot all richness ~ response diversity summaries
png('figures/response_diversity_Figs/Richness_effect_all.png', res = 600, width = 5000, height = 1750)
overall_plot + 
  xlab('Temperature strata') + 
  ylab(NULL) + 
  theme(legend.position = 'right') + 
  scale_fill_gradient2(name = bquote(cor.))
dev.off()

### Phosphorous effect
png('figures/response_diversity_Figs/Phos_max_depth_response.png', res = 600, width = 4000, height = 4000)
phos_plot + my_custom_theme_V2()
dev.off()

```

