---
title: "05_species_contributions"
author: "Conor Waldock"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

# packages relevant for data analysis
pacman::p_load(
    data.table,
  tidyverse,
  here,
  readr,
  readxl,
  gratia,
  mgcv,
  forcats,
  ggpubr,
  RColorBrewer,
  broom,
  grid,
  DHARMa,
  mgcViz,
  gridExtra,
  viridis, 
  modelsummary,
  gratia, 
  broom,
  tidyverse,
  DHARMa,
  mgcv,
  gratia,
  mgcViz
)

```


## Set up

Files are produced in earlier scripts and some of the code is copied across for similar purposes, especially from script 3 used to calculate derivatives and response diversity metrics. 

```{r}

# Read in RD functions used
source('scripts/functions/functions.R')

# Read in final ffd data
all_ffd_derivs <- readRDS('data/final_ffd.RDS')

# Bound extreme values as before
all_ffd_derivs <- all_ffd_derivs %>% 
  filter(k == 3, 
         including_electro == T) %>% 
  mutate(ffd_per_0.1degree = case_when(ffd_per_0.1degree < -0.1 ~ -0.1, 
                                       ffd_per_0.1degree > 0.1 ~ 0.1, 
                                       T ~ ffd_per_0.1degree))

# Read in species occurrences across lakes
species_occ_lakes <-  read_csv("data/taxonomy_May2025/species_lake_list_raw_EDITED.csv") %>% 
  select(Species_ecotype, Lake, Swisswide_biogeographic_status) %>% 
  unique() %>%
  rename(Species = Species_ecotype)

```

## Calculate species level contributions at each temperature bin per lake

First off establish the function and the procedure per one single lake and temperature bin

```{r}

install.packages("progress")
library(progress)


# single lake
lake <- unique(species_occ_lakes$Lake)[10]

run_sp_contributions <- lapply(unique(species_occ_lakes$Lake), function(lake){
  
  # get species present in lakes
  species_lakes <- species_occ_lakes %>% 
    filter(Lake == lake) %>% 
    pull(Species) %>% 
    unique()
  
  # get all temperatures to loop through
  all_temps <- all_ffd_derivs %>% 
      pull(mean_last_7days) %>% 
      unique() %>% 
      sort() 
  
    # Initialize progress bar for temps
    pb <- progress_bar$new(
      format = "  Processing lake :lake [:bar] :percent ETA: :eta",
      total = length(all_temps),
      clear = FALSE,
      width = 60
    )

  
  # LOOP OVER TEMPERATURE SUBSETS
  lapply(all_temps, function(temp){
    
   # update progress
   pb$tick(tokens = list(lake = lake))
    
      temp_subset <- all_ffd_derivs %>% filter(mean_last_7days == temp)
      
      if (sum(species_lakes %in% temp_subset$name) <= 2) return(NULL)
  
      # filter thermal responses to species present in different lakes
      wide_lake_ffd <- temp_subset %>% 
        ungroup() %>% 
        filter(name %in% species_lakes) %>% 
        # filter(colour == 'inside') %>% 
        select(name, mean_last_7days, ffd_per_0.1degree, k, including_electro) %>% 
        unique() %>% 
        # pivot to be wider
        pivot_wider(names_from = name, values_from = ffd_per_0.1degree)
  
      # species in lake strata
      species_lakes_strata <- species_lakes[species_lakes %in% names(wide_lake_ffd)]
  
       # loop through dropping one species at a time
       # Define species columns, assuming 'species_lakes' is a character vector of column names
       species_cols <- species_lakes_strata
       
       if(length(species_cols) <= 2) return(NULL)
       
       # LOOP OVER EACH SPECIES WITHIN TEMP SUBSET
       lapply(species_cols, function(x){
         
          sp_dropped <- species_cols[which(!species_cols %in% x)]
            
          # Convert to data.table
         wide_lake_dt <- as.data.table(wide_lake_ffd)
          
          # Perform the grouped operation
          result <- wide_lake_dt[, .(n_sp = length(na.omit(as.numeric(unlist(.SD))))+1,
                                     dissimilarity_ross = resp_div(na.omit(as.numeric(unlist(.SD))), sign_sens = F), 
                                     divergence_ross    = resp_div(na.omit(as.numeric(unlist(.SD))), sign_sens = T),
                                     mean_ffd = mean(na.omit(as.numeric(unlist(.SD)))),
                                     sd_ffd = sd(na.omit(as.numeric(unlist(.SD))))
                                     ), 
                                  by = .(k, including_electro, mean_last_7days), 
                                  .SDcols = sp_dropped]
          
          # Add lake column and the species_dropped column
          result[, c("lake", "species_dropped") := .(lake, x)]
          
          result
          
          }) %>% 
         
         # bind outputs
         bind_rows
       
       }) %>% 
    
    # capture results
    bind_rows
  
  })

# bind species contributions
sp_con_rd <- bind_rows(run_sp_contributions)

saveRDS(sp_con_rd, 'data/species_contributions_respdiv.rds')

```

Next we need to combine the calculations of response diversity with each species removed to the mean response diversity when ANY given species is removed

```{r}

# Load the data.table containing species contributions
sp_con_rd <- readRDS('data/species_contributions_respdiv.rds')

# Define the columns that you want to work with (the response metrics)
rd_cols <- c('dissimilarity_ross', 'divergence_ross', 'mean_ffd', 'sd_ffd')

# Step 1: Compute group-wise means for each of the response columns
# Grouping is done by 'k', 'including_electro', 'mean_last_7days', and 'lake'
# New columns like 'mean_dissimilarity_ross' will be created
sp_con_rd[ , 
           paste0("mean_", rd_cols) := lapply(.SD, mean),
           by = .(k, including_electro, mean_last_7days, lake), 
           .SDcols = rd_cols]

# Step 2: Subtract group means from original values for each response column
# Resulting columns will be named like 'dissimilarity_ross_diff'
# This calculates the deviation from the group mean (i.e., centered values)
sp_con_rd[, 
          paste0(rd_cols, "_diff") := Map(`-`, 
                                          .SD[, rd_cols, with = FALSE],
                                          .SD[, paste0("mean_", rd_cols), with = FALSE]), 
          .SDcols = c(rd_cols, paste0('mean_', rd_cols))]

```


# Contributions per lake overall 

Next make some plots summarising how the variation in change in species contributions looks across different lakes.

```{r}

# convert to dplyr for more intuitive code
sp_lake_rd_agg <- sp_con_rd %>% 
  as_tibble() %>% 
  group_by(species_dropped, lake) %>% 
  summarise(across(dissimilarity_ross_diff:sd_ffd_diff, c(meanLS = mean, sdLS = sd)))
  
test <- c('sd_ffd_diff_meanLS')

lake_species_heatmap <- function(var, data_set){
  
  # limits
  lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.05, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.95, na.rm = T))))
  
  ggplot(data = data_set %>%
  group_by(species_dropped) %>%
  mutate(
    species_mean = mean(get(var), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    species_dropped = fct_reorder(species_dropped, species_mean)
  ) %>%
  select(-species_mean)) + 
  geom_tile(aes(y = species_dropped, x = lake, fill = get(var))) + 
  scale_fill_gradient2(name = var,
                         low = "blue", mid = "gray50", high = "red",
                        midpoint = 0, limits = c(-lim, lim), oob = scales::squish) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ylab(NULL) + 
    xlab(NULL)
  }


vars <- c('dissimilarity_ross_diff_meanLS', 'dissimilarity_ross_diff_sdLS', 'divergence_ross_diff_meanLS', 'divergence_ross_diff_sdLS', 
  'mean_ffd_diff_meanLS', 'mean_ffd_diff_sdLS', 'sd_ffd_diff_meanLS', 'sd_ffd_diff_sdLS')

lapply(vars, function(x) lake_species_heatmap(x, sp_lake_rd_agg))


```



# Expressing difference relative to when species is included

We can also express the species contribution based on the community when it is include vs. excluded. 

```{r}

dir.create('figures/species_contributions/', recursive = T)

# Read in the original response diversity information
response_diversity <- readRDS('data/response_diversity.rds')

# rename 
response_diversity <- response_diversity %>% 
  rename(diss_com = dissimilarity_ross, 
         div_com = divergence_ross, 
         mean_com = mean_ffd, 
         sd_com = sd_ffd)

# Load the data.table containing species contributions
sp_con_rd <- readRDS('data/species_contributions_respdiv.rds')

# get response diversity contributions compared with including species
contributions_net <- left_join(sp_con_rd, response_diversity) %>% 
  filter(!is.na(div_com)) %>% 
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble() %>% 
  group_by(species_dropped, lake) %>% 
  summarise(across(diss_contr:sd_contr, c(meanLS = mean, sdLS = sd)))

# check out heatmaps for these new variables
mean_contr_plots <- lapply(c('diss_contr_meanLS', 'div_contr_meanLS', 'mean_contr_meanLS', 'sd_contr_meanLS'), 
       function(x) lake_species_heatmap(x, contributions_net))

png('figures/species_contributions/diss_allDepth.png', res = 300, height = 3500, width = 3500)
mean_contr_plots[[1]] + theme(aspect.ratio = 1.5)
dev.off()

png('figures/species_contributions/div_allDepth.png', res = 300, height = 3500, width = 3500)
mean_contr_plots[[2]] + theme(aspect.ratio = 1)
dev.off()

png('figures/species_contributions/mean_allDepth.png', res = 300, height = 3500, width = 3500)
mean_contr_plots[[3]] + theme(aspect.ratio = 1)
dev.off()

png('figures/species_contributions/sd_allDepth.png', res = 300, height = 3500, width = 3500)
mean_contr_plots[[4]] + theme(aspect.ratio = 1)
dev.off()


sd_contr_plots <- lapply(c('diss_contr_sdLS', 'div_contr_sdLS', 'mean_contr_sdLS', 'sd_contr_sdLS'), 
       function(x) lake_species_heatmap(x, contributions_net))

```


# Assess patterns across temperature as in other analyses

```{r}

# join in raw response diversities and species contributions and aggregate across temperature bins
contributions_net_tempbin <- left_join(sp_con_rd, response_diversity) %>% 
  # remove communities without estimates
  filter(!is.na(div_com)) %>% 
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble() %>% 
  # Create temperature bins 
  mutate(temp_bin = paste0('temp_', sprintf("%02d", plyr::round_any(mean_last_7days, accuracy = 5)))) %>% 
  # Group the data by the 'lake' column
  group_by(species_dropped, lake, temp_bin) %>% 
  # For all columns from 'dissimilarity_ross' to 'sd_ffd', compute the mean
  # 'na.rm = T' ensures that NA values are ignored during the mean calculation
  summarise(across(diss_contr:sd_contr, c(meanLSD = mean, sdLSD = sd))) %>% 
  filter(temp_bin != 'temp_25')


# create lake species heatmaps per depthbin

lake_species_depth_heatmap <- function(var, data_set){
  
  # limits
  lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.05, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.95, na.rm = T))))
  
  ggplot(data = data_set %>%
  group_by(species_dropped) %>%
  mutate(
    species_mean = mean(get(var), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    species_dropped = fct_reorder(species_dropped, species_mean)
  ) %>%
  select(-species_mean)) + 
  geom_tile(aes(y = species_dropped, x = lake, fill = get(var))) + 
  scale_fill_gradient2(name = var,
                         low = "blue", mid = "gray50", high = "red",
                        midpoint = 0, limits = c(-lim, lim), oob = scales::squish) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ylab(NULL) + 
    xlab(NULL)  +
    facet_wrap(~temp_bin)
}

mean_contr_tempbin_plots <- lapply(c('diss_contr_meanLSD', 'div_contr_meanLSD', 'mean_contr_meanLSD', 'sd_contr_meanLSD'), 
       function(x) lake_species_depth_heatmap(x, contributions_net_tempbin))


png('figures/species_contributions/diss_tempBin.png', res = 300, height = 4500, width = 4500)
mean_contr_tempbin_plots[[1]] 
dev.off()

png('figures/species_contributions/div_tempBin.png', res = 300, height = 4500, width = 4500)
mean_contr_tempbin_plots[[2]] 
dev.off()

png('figures/species_contributions/mean_tempBin.png', res = 300, height = 4500, width = 4500)
mean_contr_tempbin_plots[[3]] 
dev.off()

png('figures/species_contributions/sd_tempBin.png', res = 300, height = 4500, width = 4500)
mean_contr_tempbin_plots[[4]] 
dev.off()

```



# Aggregate species response contributions across lakes but keep thermal structure

```{r}

mean_contr_tempbin_nolake_plots <- left_join(sp_con_rd, response_diversity) %>% 
  # remove communities without estimates
  filter(!is.na(div_com)) %>% 
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble() %>% 
  # Create temperature bins 
  mutate(temp_bin = paste0('temp_', sprintf("%02d", plyr::round_any(mean_last_7days, accuracy = 5)))) %>% 
  # Group the data by the 'lake' column
  group_by(species_dropped, temp_bin) %>% 
  # For all columns from 'dissimilarity_ross' to 'sd_ffd', compute the mean
  # 'na.rm = T' ensures that NA values are ignored during the mean calculation
  summarise(across(diss_contr:sd_contr, c(meanLSD = ~ mean(.x, na.rm = TRUE), sdLSD = ~ sd(.x, na.rm = TRUE)))) %>% 
  filter(temp_bin != 'temp_25')


species_depth_heatmap <- function(var, data_set){
  
  # limits
  lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.05, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.95, na.rm = T))))
  
  ggplot(data = data_set %>%
  group_by(species_dropped) %>%
  mutate(
    species_mean = mean(get(var), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    species_dropped = fct_reorder(species_dropped, species_mean)
  ) %>%
  select(-species_mean)) + 
    
  geom_tile(aes(y = species_dropped, x = temp_bin, fill = get(var))) + 
  scale_fill_gradient2(name = var,
                         low = "blue", mid = "gray50", high = "red",
                        midpoint = 0, limits = c(-lim, lim), oob = scales::squish) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ylab(NULL) + 
    xlab(NULL) 
}

species_tempbin_agg_plots_mean <- lapply(c('diss_contr_meanLSD', 'div_contr_meanLSD', 'mean_contr_meanLSD', 'sd_contr_meanLSD'), 
       function(x) species_depth_heatmap(x, mean_contr_tempbin_nolake_plots))


png('figures/species_contributions/diss_tempBin_agg.png', res = 300, height = 4500, width = 2000)
species_tempbin_agg_plots_mean[[1]] 
dev.off()

png('figures/species_contributions/div_tempBin_agg.png', res = 300, height = 4500, width = 2000)
species_tempbin_agg_plots_mean[[2]] 
dev.off()

png('figures/species_contributions/mean_tempBin_agg.png', res = 300, height = 4500, width = 2000)
species_tempbin_agg_plots_mean[[3]] 
dev.off()

png('figures/species_contributions/sd_tempBin_agg.png', res = 300, height = 4500, width = 2000)
species_tempbin_agg_plots_mean[[4]] 
dev.off()


species_tempbin_agg_plots_sd <- lapply(c('diss_contr_sdLSD', 'div_contr_sdLSD', 'mean_contr_sdLSD', 'sd_contr_sdLSD'), 
       function(x) species_depth_heatmap(x, mean_contr_tempbin_nolake_plots))



```

