---
title: "05_species_contributions"
author: "Conor Waldock"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

# packages relevant for data analysis
pacman::p_load(
    data.table,
  tidyverse,
  here,
  readr,
  readxl,
  gratia,
  mgcv,
  forcats,
  ggpubr,
  RColorBrewer,
  broom,
  grid,
  DHARMa,
  mgcViz,
  gridExtra,
  viridis, 
  modelsummary,
  gratia, 
  broom,
  tidyverse,
  DHARMa,
  mgcv,
  gratia,
  mgcViz
)

```


## Set up

Files are produced in earlier scripts and some of the code is copied across for similar purposes, especially from script 3 used to calculate derivatives and response diversity metrics. 

```{r}

# Read in RD functions used
source('scripts/functions/functions.R')

# Read in final ffd data
all_ffd_derivs <- readRDS('data/final_ffd.RDS')

# Bound extreme values as before
all_ffd_derivs <- all_ffd_derivs %>% 
  filter(k == 3, 
         including_electro == T) %>% 
  mutate(ffd_per_0.1degree = case_when(ffd_per_0.1degree < -0.1 ~ -0.1, 
                                       ffd_per_0.1degree > 0.1 ~ 0.1, 
                                       T ~ ffd_per_0.1degree))

# Read in species occurrences across lakes
species_occ_lakes <- read_csv("data/taxonomy_May2025/species_lake_list_raw_EDITED.csv") %>% 
  select(Species_ecotype, Lake, Swisswide_biogeographic_status) %>% 
  unique() %>%
  rename(Species = Species_ecotype)

```

## Format swiss biogeographic categorisation for saving as a table
```{r}

# Clean species names as in figures
species_occ_lakes_v2 <- read_csv("data/taxonomy_May2025/species_lake_list_raw_EDITED.csv")

# format for presenting in wide tale
id_table <- species_occ_lakes_v2 %>% 
  
  # get columns of focus
  dplyr::select(Species_ecotype, Lake, Lake_biogeographic_status) %>% 
  filter(Lake_biogeographic_status != 'remove') %>% 
  unique() %>%
  rename(Species = Species_ecotype) %>% 

  # revalue the levels of temperature bins
  mutate(status = dplyr::recode(Lake_biogeographic_status, 
                                       'native' = 'WN',
                                       'endemic' = 'ISD',
                                       'non_native' = 'NN'), 
         Species = gsub('_', ' ', Species), 
         fill = 'x') %>% 
  
  # remove old coding
  select(-Lake_biogeographic_status) %>% 
  
  # pivot to shorten
  pivot_wider(names_from = Lake, values_from = fill, values_fill = ' ')

write_csv(id_table, 'figures/species_contributions/species_lake_table.csv')
  
```


## Calculate species level contributions at each temperature bin per lake

First off establish the function and the procedure per one single lake and temperature bin

```{r}

install.packages("progress")
library(progress)


# single lake
lake <- unique(species_occ_lakes$Lake)[10]

run_sp_contributions <- lapply(unique(species_occ_lakes$Lake), function(lake){
  
  # get species present in lakes
  species_lakes <- species_occ_lakes %>% 
    filter(Lake == lake) %>% 
    pull(Species) %>% 
    unique()
  
  # get all temperatures to loop through
  all_temps <- all_ffd_derivs %>% 
      pull(mean_last_7days) %>% 
      unique() %>% 
      sort() 
  
    # Initialize progress bar for temps
    pb <- progress_bar$new(
      format = "  Processing lake :lake [:bar] :percent ETA: :eta",
      total = length(all_temps),
      clear = FALSE,
      width = 60
    )

  
  # LOOP OVER TEMPERATURE SUBSETS
  lapply(all_temps, function(temp){
    
   # update progress
   pb$tick(tokens = list(lake = lake))
    
      temp_subset <- all_ffd_derivs %>% filter(mean_last_7days == temp)
      
      if (sum(species_lakes %in% temp_subset$name) <= 2) return(NULL)
  
      # filter thermal responses to species present in different lakes
      wide_lake_ffd <- temp_subset %>% 
        ungroup() %>% 
        filter(name %in% species_lakes) %>% 
        # filter(colour == 'inside') %>% 
        select(name, mean_last_7days, ffd_per_0.1degree, k, including_electro) %>% 
        unique() %>% 
        # pivot to be wider
        pivot_wider(names_from = name, values_from = ffd_per_0.1degree)
  
      # species in lake strata
      species_lakes_strata <- species_lakes[species_lakes %in% names(wide_lake_ffd)]
  
       # loop through dropping one species at a time
       # Define species columns, assuming 'species_lakes' is a character vector of column names
       species_cols <- species_lakes_strata
       
       if(length(species_cols) <= 2) return(NULL)
       
       # LOOP OVER EACH SPECIES WITHIN TEMP SUBSET
       lapply(species_cols, function(x){
         
          sp_dropped <- species_cols[which(!species_cols %in% x)]
            
          # Convert to data.table
         wide_lake_dt <- as.data.table(wide_lake_ffd)
          
          # Perform the grouped operation
          result <- wide_lake_dt[, .(n_sp = length(na.omit(as.numeric(unlist(.SD))))+1,
                                     dissimilarity_ross = resp_div(na.omit(as.numeric(unlist(.SD))), sign_sens = F), 
                                     divergence_ross    = resp_div(na.omit(as.numeric(unlist(.SD))), sign_sens = T),
                                     mean_ffd = mean(na.omit(as.numeric(unlist(.SD)))),
                                     sd_ffd = sd(na.omit(as.numeric(unlist(.SD))))
                                     ), 
                                  by = .(k, including_electro, mean_last_7days), 
                                  .SDcols = sp_dropped]
          
          # Add lake column and the species_dropped column
          result[, c("lake", "species_dropped") := .(lake, x)]
          
          result
          
          }) %>% 
         
         # bind outputs
         bind_rows
       
       }) %>% 
    
    # capture results
    bind_rows
  
  })

# bind species contributions
sp_con_rd <- bind_rows(run_sp_contributions)

saveRDS(sp_con_rd, 'data/species_contributions_respdiv.rds')

```

Next we need to combine the calculations of response diversity with each species removed to the mean response diversity when ANY given species is removed

```{r}

# Load the data.table containing species contributions
sp_con_rd <- readRDS('data/species_contributions_respdiv.rds')

# Define the columns that you want to work with (the response metrics)
rd_cols <- c('dissimilarity_ross', 'divergence_ross', 'mean_ffd', 'sd_ffd')

# Step 1: Compute group-wise means for each of the response columns
# Grouping is done by 'k', 'including_electro', 'mean_last_7days', and 'lake'
# New columns like 'mean_dissimilarity_ross' will be created
sp_con_rd[ , 
           paste0("mean_", rd_cols) := lapply(.SD, mean),
           by = .(k, including_electro, mean_last_7days, lake), 
           .SDcols = rd_cols]

# Step 2: Subtract group means from original values for each response column
# Resulting columns will be named like 'dissimilarity_ross_diff'
# This calculates the deviation from the group mean (i.e., centered values)
sp_con_rd[, 
          paste0(rd_cols, "_diff") := Map(`-`, 
                                          .SD[, rd_cols, with = FALSE],
                                          .SD[, paste0("mean_", rd_cols), with = FALSE]), 
          .SDcols = c(rd_cols, paste0('mean_', rd_cols))]

```


# Contributions per lake overall 

Next make some plots summarising how the variation in change in species contributions looks across different lakes.

```{r}

# convert to dplyr for more intuitive code
sp_lake_rd_agg <- sp_con_rd %>% 
  as_tibble() %>% 
  group_by(species_dropped, lake) %>% 
  summarise(across(dissimilarity_ross_diff:sd_ffd_diff, c(meanLS = mean, sdLS = sd)))
  
test <- c('sd_ffd_diff_meanLS')

lake_species_heatmap <- function(var, data_set){
  
  # limits
  lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.05, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.95, na.rm = T))))
  
  ggplot(data = data_set %>%
  group_by(species_dropped) %>%
  mutate(
    species_mean = mean(get(var), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    species_dropped = fct_reorder(species_dropped, species_mean)
  ) %>%
  select(-species_mean)) + 
  geom_tile(aes(y = species_dropped, x = lake, fill = get(var))) + 
  scale_fill_gradient2(name = var,
                         
                        midpoint = 0, limits = c(-lim, lim), oob = scales::squish) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ylab(NULL) + 
    xlab(NULL)
  }


vars <- c('dissimilarity_ross_diff_meanLS', 'dissimilarity_ross_diff_sdLS', 'divergence_ross_diff_meanLS', 'divergence_ross_diff_sdLS', 
  'mean_ffd_diff_meanLS', 'mean_ffd_diff_sdLS', 'sd_ffd_diff_meanLS', 'sd_ffd_diff_sdLS')

lapply(vars, function(x) lake_species_heatmap(x, sp_lake_rd_agg))


```



# Expressing difference relative to when species is included

We can also express the species contribution based on the community when it is include vs. excluded. 

```{r}

dir.create('figures/species_contributions/', recursive = T)

# Read in the original response diversity information
response_diversity <- readRDS('data/response_diversity.rds')

# rename 
response_diversity <- response_diversity %>% 
  rename(diss_com = dissimilarity_ross, 
         div_com = divergence_ross, 
         mean_com = mean_ffd, 
         sd_com = sd_ffd)

# Load the data.table containing species contributions
sp_con_rd <- readRDS('data/species_contributions_respdiv.rds')

# get response diversity contributions compared with including species
contributions_net <- left_join(sp_con_rd, response_diversity) %>% 
  filter(!is.na(div_com)) %>% 
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble() %>% 
  group_by(species_dropped, lake) %>% 
  summarise(across(diss_contr:sd_contr, c(meanLS = mean, sdLS = sd)))

# check out heatmaps for these new variables
mean_contr_plots <- lapply(c('diss_contr_meanLS', 'div_contr_meanLS', 'mean_contr_meanLS', 'sd_contr_meanLS'), 
       function(x) lake_species_heatmap(x, contributions_net))

png('figures/species_contributions/diss_allDepth.png', res = 300, height = 3500, width = 3500)
mean_contr_plots[[1]] + theme(aspect.ratio = 1.5)
dev.off()

png('figures/species_contributions/div_allDepth.png', res = 300, height = 3500, width = 3500)
mean_contr_plots[[2]] + theme(aspect.ratio = 1)
dev.off()

png('figures/species_contributions/mean_allDepth.png', res = 300, height = 3500, width = 3500)
mean_contr_plots[[3]] + theme(aspect.ratio = 1)
dev.off()

png('figures/species_contributions/sd_allDepth.png', res = 300, height = 3500, width = 3500)
mean_contr_plots[[4]] + theme(aspect.ratio = 1)
dev.off()


sd_contr_plots <- lapply(c('diss_contr_sdLS', 'div_contr_sdLS', 'mean_contr_sdLS', 'sd_contr_sdLS'), 
       function(x) lake_species_heatmap(x, contributions_net))

```


# Assess patterns across temperature as in other analyses

```{r}

# join in raw response diversities and species contributions and aggregate across temperature bins
contributions_net_tempbin <- left_join(sp_con_rd, response_diversity) %>% 
  # remove communities without estimates
  filter(!is.na(div_com)) %>% 
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble() %>% 
  # Create temperature bins 
  mutate(temp_bin = paste0('temp_', sprintf("%02d", plyr::round_any(mean_last_7days, accuracy = 5)))) %>% 
  # Group the data by the 'lake' column
  group_by(species_dropped, lake, temp_bin) %>% 
  # For all columns from 'dissimilarity_ross' to 'sd_ffd', compute the mean
  # 'na.rm = T' ensures that NA values are ignored during the mean calculation
  summarise(across(diss_contr:sd_contr, c(meanLSD = mean, sdLSD = sd))) %>% 
  filter(temp_bin != 'temp_25')


# create lake species heatmaps per depthbin

lake_species_depth_heatmap <- function(var, data_set){
  
  # limits
  lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.05, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.95, na.rm = T))))
  
  ggplot(data = data_set %>%
  group_by(species_dropped) %>%
  mutate(
    species_mean = mean(get(var), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    species_dropped = fct_reorder(species_dropped, species_mean)
  ) %>%
  select(-species_mean)) + 
  geom_tile(aes(y = species_dropped, x = lake, fill = get(var))) + 
  scale_fill_gradient2(name = var,
                        # low = "blue", mid = "gray50", high = "red",
                        midpoint = 0, limits = c(-lim, lim), oob = scales::squish) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ylab(NULL) + 
    xlab(NULL)  +
    facet_wrap(~temp_bin)
}

mean_contr_tempbin_plots <- lapply(c('diss_contr_meanLSD', 'div_contr_meanLSD', 'mean_contr_meanLSD', 'sd_contr_meanLSD'), 
       function(x) lake_species_depth_heatmap(x, contributions_net_tempbin))


png('figures/species_contributions/diss_tempBin.png', res = 300, height = 4500, width = 4500)
mean_contr_tempbin_plots[[1]] 
dev.off()

png('figures/species_contributions/div_tempBin.png', res = 300, height = 4500, width = 4500)
mean_contr_tempbin_plots[[2]] 
dev.off()

png('figures/species_contributions/mean_tempBin.png', res = 300, height = 4500, width = 4500)
mean_contr_tempbin_plots[[3]] 
dev.off()

png('figures/species_contributions/sd_tempBin.png', res = 300, height = 4500, width = 4500)
mean_contr_tempbin_plots[[4]] 
dev.off()

```



# Aggregate species response contributions across lakes but keep thermal structure

```{r}

mean_contr_tempbin_nolake_plots <- left_join(sp_con_rd, response_diversity) %>% 
  # remove communities without estimates
  filter(!is.na(div_com)) %>% 
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble() %>% 
  # Create temperature bins 
  mutate(temp_bin = paste0('temp_', sprintf("%02d", plyr::round_any(mean_last_7days, accuracy = 5)))) %>% 
  # Group the data by the 'lake' column
  group_by(species_dropped, temp_bin) %>% 
  # For all columns from 'dissimilarity_ross' to 'sd_ffd', compute the mean
  # 'na.rm = T' ensures that NA values are ignored during the mean calculation
  summarise(across(diss_contr:sd_contr, c(meanLSD = ~ mean(.x, na.rm = TRUE), sdLSD = ~ sd(.x, na.rm = TRUE)))) %>% 
  filter(temp_bin != 'temp_25')


species_depth_heatmap <- function(var, data_set){
  
  # limits
  lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.05, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.95, na.rm = T))))
  
  ggplot(data = data_set %>%
  group_by(species_dropped) %>%
  mutate(
    species_mean = mean(get(var), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    species_dropped = fct_reorder(species_dropped, species_mean)
  ) %>%
  select(-species_mean)) + 
    
  geom_tile(aes(y = species_dropped, x = temp_bin, fill = get(var))) + 
  scale_fill_gradient2(name = var,
                         low = "blue", mid = "gray50", high = "red",
                        midpoint = 0, limits = c(-lim, lim), oob = scales::squish) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ylab(NULL) + 
    xlab(NULL) 
}

species_tempbin_agg_plots_mean <- lapply(c('diss_contr_meanLSD', 'div_contr_meanLSD', 'mean_contr_meanLSD', 'sd_contr_meanLSD'), 
       function(x) species_depth_heatmap(x, mean_contr_tempbin_nolake_plots))

png('figures/species_contributions/diss_tempBin_agg.png', res = 300, height = 4500, width = 2000)
species_tempbin_agg_plots_mean[[1]] 
dev.off()

png('figures/species_contributions/div_tempBin_agg.png', res = 300, height = 4500, width = 2000)
species_tempbin_agg_plots_mean[[2]] 
dev.off()

png('figures/species_contributions/mean_tempBin_agg.png', res = 300, height = 4500, width = 2000)
species_tempbin_agg_plots_mean[[3]] 
dev.off()

png('figures/species_contributions/sd_tempBin_agg.png', res = 300, height = 4500, width = 2000)
species_tempbin_agg_plots_mean[[4]] 
dev.off()


species_tempbin_agg_plots_sd <- lapply(c('diss_contr_sdLSD', 'div_contr_sdLSD', 'mean_contr_sdLSD', 'sd_contr_sdLSD'), 
       function(x) species_depth_heatmap(x, mean_contr_tempbin_nolake_plots))

```


# Join in biogeographic realms of species and plot with faceted realms

```{r}

# process biogeographic data per lake
species_biog_lake <- read_csv("data/taxonomy_May2025/species_lake_list_raw_EDITED.csv") %>% 
  select(Species_ecotype, Lake, Lake_biogeographic_status) %>% 
  unique() %>%
  rename(Species = Species_ecotype) %>% 
  filter(Lake_biogeographic_status != 'remove', 
         !is.na(Lake_biogeographic_status))

# aggregation of response contributions over biogeographic status
contri_biog_data <- 
  
  # join together species contributions and overall community response diversity
  left_join(sp_con_rd, response_diversity) %>%
  
  # join in biogeographic status at species and lake scale
  left_join(., species_biog_lake %>% unique(), by = c('species_dropped' = 'Species', 'lake' = 'Lake')) %>% 
  
  # remove communities without estimates
  filter(!is.na(div_com), !is.na(Lake_biogeographic_status)) %>% 
  
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble() %>% 
  
  # Create temperature bins 
  mutate(temp_bin = paste0('temp_', sprintf("%02d", plyr::round_any(mean_last_7days, accuracy = 5)))) %>% 
  
  # Group the data by the 'lake' column
  group_by(species_dropped, temp_bin) %>% 
  
  # For all columns from 'dissimilarity_ross' to 'sd_ffd', compute the mean
  # 'na.rm = T' ensures that NA values are ignored during the mean calculation
  summarise(across(diss_contr:sd_contr, c(meanLSDB = ~ mean(.x, na.rm = TRUE), sdLSDB = ~ sd(.x, na.rm = TRUE)))) %>% 
  
  # Remove temperature bin with very few species
  filter(temp_bin != 'temp_25') %>% 
  
  # join in swiss wide status
  left_join(., species_occ_lakes %>% select(-Lake) %>% unique(), by = c('species_dropped' = 'Species')) %>% 
  
  # remove underscores from species names
  mutate(species_dropped = gsub('_', ' ', species_dropped)) %>% 
  
  # revalue the levels of temperature bins
  mutate(
         temp_bin  = factor(dplyr::recode(temp_bin, 
                                   'temp_05' = '2.5 to 7.5°C',
                                   'temp_10' = '7.5 to 12.5°C',
                                   'temp_15' = '12.5 to 17.5°C',
                                   'temp_20' = '17.5 to 22.5°C'),
                            levels = c('2.5 to 7.5°C', '7.5 to 12.5°C', '12.5 to 17.5°C', '17.5 to 22.5°C')))


## Create funtion to plot
pacman::p_load(ggtext)  # enables colored axis labels

species_depth_biog_heatmap <- function(var, data_set){
  
  # limits
  if(var == 'div_contr_meanLSDB'){lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.1, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.9, na.rm = T))))}else{
    
    lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.05, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.95, na.rm = T))))}
  
  # Step 1: reorder species_dropped based on species_mean
data_set <- data_set %>%
  group_by(species_dropped) %>%
  mutate(species_mean = mean(get(var), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    species_dropped = fct_reorder(species_dropped, species_mean)
  )

# Step 2: create species_label and apply the same factor levels
data_set <- data_set %>%
  mutate(
    status_symbol = case_when(
      Swisswide_biogeographic_status == "endemic" ~ "&#9733;",      # ★
      Swisswide_biogeographic_status == "non_native" ~ "&#9830;",   # ♦
      TRUE ~ "--"  # No symbol for others
    ),
    species_label_text = paste0(
      "<span style='color:",
      ifelse(.data$Swisswide_biogeographic_status == "endemic", "#3CB371",
      ifelse(.data$Swisswide_biogeographic_status == "non_native", "purple", "#C7C35E")),
      "'>", species_dropped, '  ', status_symbol, "</span>"
    ),
    # 🧠 Make this a factor with the same order as species_dropped
    species_label = factor(species_label_text, levels = unique(species_label_text[order(species_dropped)]))
  )

  
  ggplot(data_set) + 
    geom_tile(aes(y = species_label, x = temp_bin, fill = get(var))) + 
    scale_fill_gradient2(
      name = NULL, 
      mid = "gray90",
      midpoint = 0, limits = c(-lim, lim), oob = scales::squish
    ) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.ticks.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(hjust = 0),
      
      # ✅ enable rich text formatting for y-axis
      axis.text.y = element_markdown()
    ) +
    ylab(NULL) +
    xlab(NULL)
}

contri_biog_plots <- lapply(c('diss_contr_meanLSDB', 'div_contr_meanLSDB', 
                                           'mean_contr_meanLSDB', 'sd_contr_meanLSDB'), 
       function(x) species_depth_biog_heatmap(x, contri_biog_data))

contri_biog_plots[[1]]
contri_biog_plots[[3]]

contri_biog_plots[[2]] + ggtitle('(a) Divergence')  +
contri_biog_plots[[4]] + ggtitle('(b) SD response')

png('figures/species_contributions/spContributions_biog_lake_temp_contributions.png', res = 300, height = 2300, width = 4000)
contri_biog_plots[[4]] + ggtitle('(a) response SD') + 
contri_biog_plots[[2]] + ggtitle('(b) response divergence')
dev.off()


png('figures/species_contributions/spContributions_biog_lake_temp_contributions_ALL.png', res = 300, height = 5000, width = 4000)
contri_biog_plots[[1]] + ggtitle('(a) response dissimilarity') + 
  contri_biog_plots[[4]] + ggtitle('(b) response SD') + 
    contri_biog_plots[[2]] + ggtitle('(c) response divergence') + 
contri_biog_plots[[3]] + ggtitle('(d) response direction')
dev.off()

png('figures/species_contributions/spContributions_lake_temp_contributions_SD.png', res = 300, height = 2000, width = 1700)
contri_biog_plots[[4]] + theme(aspect.ratio = 4)
dev.off()

png('figures/species_contributions/spContributions_lake_temp_contributions_Div.png', res = 300, height = 2000, width = 1700)
contri_biog_plots[[2]]+ theme(aspect.ratio = 4)
dev.off()

png('figures/species_contributions/spContributions_lake_temp_contributions_Dir.png', res = 300, height = 2000, width = 1700)
contri_biog_plots[[3]]+ theme(aspect.ratio = 4)
dev.off()


```

# Join in biogeographic realms of species and plot with faceted realms

```{r}

# process biogeographic data per lake
species_biog_lake <- read_csv("data/taxonomy_May2025/species_lake_list_raw_EDITED.csv") %>% 
  select(Species_ecotype, Lake, Lake_biogeographic_status) %>% 
  unique() %>%
  rename(Species = Species_ecotype) %>% 
  filter(Lake_biogeographic_status != 'remove', 
         !is.na(Lake_biogeographic_status))

# aggregation of response contributions over biogeographic status
contri_biog_data <- 
  
  # join together species contributions and overall community response diversity
  left_join(sp_con_rd, response_diversity) %>%
  
  # join in biogeographic status at species and lake scale
  left_join(., species_biog_lake %>% unique(), by = c('species_dropped' = 'Species', 'lake' = 'Lake')) %>% 
  
  # remove communities without estimates
  filter(!is.na(div_com), !is.na(Lake_biogeographic_status)) %>% 
  
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble() %>% 
  
  # Create temperature bins 
  mutate(temp_bin = paste0('temp_', sprintf("%02d", plyr::round_any(mean_last_7days, accuracy = 5)))) %>% 
  
  # Group the data by the 'lake' column
  group_by(species_dropped, Lake_biogeographic_status, temp_bin) %>% 
  
  # For all columns from 'dissimilarity_ross' to 'sd_ffd', compute the mean
  # 'na.rm = T' ensures that NA values are ignored during the mean calculation
  summarise(across(diss_contr:sd_contr, c(meanLSDB = ~ mean(.x, na.rm = TRUE), sdLSDB = ~ sd(.x, na.rm = TRUE)))) %>% 
  
  # Remove temperature bin with very few species
  filter(temp_bin != 'temp_25') %>% 
  
  # join in swiss wide status
  left_join(., species_occ_lakes %>% select(-Lake) %>% unique(), by = c('species_dropped' = 'Species')) %>% 
  
  # remove underscores from species names
  mutate(species_dropped = gsub('_', ' ', species_dropped)) %>% 
  
  # revalue the levels of temperature bins
  mutate(Lake_biogeographic_status = factor(dplyr::recode(Lake_biogeographic_status, 
                                       'native' = 'widespread \nnative',
                                       'endemic' = 'in situ \ndiversification',
                                       'non_native' = 'non-native'), levels = c('in situ \ndiversification', 
                                                                                'widespread \nnative', 
                                                                                'non-native')),
         temp_bin  = factor(dplyr::recode(temp_bin, 
                                   'temp_05' = '2.5 to 7.5°C',
                                   'temp_10' = '7.5 to 12.5°C',
                                   'temp_15' = '12.5 to 17.5°C',
                                   'temp_20' = '17.5 to 22.5°C'),
                            levels = c('2.5 to 7.5°C', '7.5 to 12.5°C', '12.5 to 17.5°C', '17.5 to 22.5°C')))


## Create funtion to plot
pacman::p_load(ggtext)  # enables colored axis labels

species_depth_biog_heatmap <- function(var, data_set){
  
  # limits
  if(var == 'div_contr_meanLSDB'){lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.1, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.9, na.rm = T))))}else{
    
    lim <- max(abs(c(data_set %>% pull(get(var)) %>% quantile(., 0.05, na.rm = T), data_set %>% pull(get(var)) %>% quantile(., 0.95, na.rm = T))))}
  
  # Step 1: reorder species_dropped based on species_mean
data_set <- data_set %>%
  group_by(species_dropped) %>%
  mutate(species_mean = mean(get(var), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    species_dropped = fct_reorder(species_dropped, species_mean)
  )

# Step 2: create species_label and apply the same factor levels
data_set <- data_set %>%
  mutate(
    status_symbol = case_when(
      Swisswide_biogeographic_status == "endemic" ~ "&#9733;",      # ★
      Swisswide_biogeographic_status == "non_native" ~ "&#9830;",   # ♦
      TRUE ~ "--"  # No symbol for others
    ),
    species_label_text = paste0(
      "<span style='color:",
      ifelse(.data$Swisswide_biogeographic_status == "endemic", "#3CB371",
      ifelse(.data$Swisswide_biogeographic_status == "non_native", "purple", "#C7C35E")),
      "'>", species_dropped, '  ', status_symbol, "</span>"
    ),
    # 🧠 Make this a factor with the same order as species_dropped
    species_label = factor(species_label_text, levels = unique(species_label_text[order(species_dropped)]))
  )

  
  ggplot(data_set) + 
    geom_tile(aes(y = species_label, x = temp_bin, fill = get(var))) + 
    facet_wrap(~Lake_biogeographic_status) + 
    scale_fill_gradient2(
      name = NULL, 
      mid = "gray90",
      midpoint = 0, limits = c(-lim, lim), oob = scales::squish
    ) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.ticks.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(hjust = 0),
      
      # ✅ enable rich text formatting for y-axis
      axis.text.y = element_markdown()
    ) +
    ylab(NULL) +
    xlab(NULL)
}

contri_biog_plots <- lapply(c('diss_contr_meanLSDB', 'div_contr_meanLSDB', 
                                           'mean_contr_meanLSDB', 'sd_contr_meanLSDB'), 
       function(x) species_depth_biog_heatmap(x, contri_biog_data))

contri_biog_plots[[1]]
contri_biog_plots[[3]]

contri_biog_plots[[2]] + ggtitle('(a) Divergence')  +
contri_biog_plots[[4]] + ggtitle('(b) SD response')

png('figures/species_contributions/spContributions_biog_lake_temp_contributions.png', res = 300, height = 2300, width = 4000)
contri_biog_plots[[4]] + ggtitle('(a) response SD') + 
contri_biog_plots[[2]] + ggtitle('(b) response divergence')
dev.off()


png('figures/species_contributions/spContributions_biog_lake_temp_contributions_ALL.png', res = 300, height = 5000, width = 4000)
contri_biog_plots[[1]] + ggtitle('(a) response dissimilarity') + 
  contri_biog_plots[[4]] + ggtitle('(b) response SD') + 
    contri_biog_plots[[2]] + ggtitle('(c) response divergence') + 
contri_biog_plots[[3]] + ggtitle('(d) response direction')
dev.off()

png('figures/species_contributions/spContributions_biog_lake_temp_contributions_SD.png', res = 300, height = 2000, width = 1700)
contri_biog_plots[[4]]
dev.off()

```


Using the dataframes created in the above chunk, rather than plotted as heatmaps investigate as summary distributions using box/beehive/violin plots.

```{r}

# focal metrics
focal_metrics <- c('diss_contr_meanLSDB', 'div_contr_meanLSDB', 'mean_contr_meanLSDB', 'sd_contr_meanLSDB')

# pivot to create long dataframe for ggplot2
contri_biog_anova <- contri_biog_data %>% 
  
  select(species_dropped:temp_bin, Swisswide_biogeographic_status, focal_metrics) %>% 
  
  pivot_longer(cols = focal_metrics) %>% 
  
  # relevel labels
  mutate(name = case_match(name, "diss_contr_meanLSDB" ~ "response dissimilarity", 
                                 "div_contr_meanLSDB"  ~ "response divergence", 
                                 "mean_contr_meanLSDB" ~ "response direction",
                                 "sd_contr_meanLSDB"   ~ "response SD")) %>% 
   mutate(name = fct_relevel(name, "response dissimilarity", "response SD", "response divergence", "response direction"))

# Create alternating shading data for background
temp_levels <- unique(contri_biog_anova$temp_bin)

background_rects <- data.frame(
  temp_bin = temp_levels,
  xmin = seq_along(temp_levels) - 0.5,
  xmax = seq_along(temp_levels) + 0.5,
  fill_bg = rep(c("gray95", "white"), length.out = length(temp_levels))
)

# Plot
ggplot(data = contri_biog_anova %>% 
         # here when contributions to response divergence == 0 remove
         filter(case_when(name == "response divergence" ~ value != 0, T~T)), 
       aes(x = temp_bin, y = value)) +

  # Background shading with separate fill aesthetic (fill_bg)
  geom_rect(data = background_rects, inherit.aes = FALSE,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = fill_bg),
            alpha = 0.5) +
  
  geom_hline(aes(yintercept = 0)) + 

  # Manual fill scale just for background
  scale_fill_manual(values = c("gray95", "white"), guide = "none") +

  # Main plot layers
  geom_boxplot(aes(colour = Lake_biogeographic_status),
               outlier.shape = NA, width = 0.5, alpha = 0.3,
               position = position_dodge(width = 0.75)) +
  
  scale_colour_manual(name = 'biogeography', values = c(c("#3CB371",  "#C7C35E", 'purple'))) +

  ggbeeswarm::geom_beeswarm(aes(color = Lake_biogeographic_status), 
                dodge.width = 0.75, cex = 1, priority = "density") +

  facet_wrap(~name, scales = "free") +
  theme_bw() +
  theme(panel.grid = element_blank(), 
        legend.position = 'none') +
  labs(x = "Temperature Bin", y = "Value")



```

# Testing whether groups are different from 0 using wilcoxon signed rank test

```{r}

# focal metrics
focal_metrics <- c('diss_contr_meanLSDB', 'div_contr_meanLSDB', 'mean_contr_meanLSDB', 'sd_contr_meanLSDB')

# pivot to create long dataframe for ggplot2
contri_biog_anova <- contri_biog_data %>% 
  
  select(species_dropped:temp_bin, Swisswide_biogeographic_status, focal_metrics) %>% 
  
  pivot_longer(cols = focal_metrics) %>% 
  
  # relevel labels
  mutate(name = case_match(name, "diss_contr_meanLSDB" ~ "response dissimilarity", 
                                 "div_contr_meanLSDB"  ~ "response divergence", 
                                 "mean_contr_meanLSDB" ~ "response direction",
                                 "sd_contr_meanLSDB"   ~ "response SD")) %>% 
   mutate(name = fct_relevel(name, "response dissimilarity", "response SD", "response divergence", "response direction"))

remove_outliers <- function(x) {
  qnt <- quantile(x, probs = c(0.25, 0.75), na.rm = TRUE)
  H <- 1.5 * IQR(x, na.rm = TRUE)
  x[x < (qnt[1] - H) | x > (qnt[2] + H)] <- NA
  return(x)
}

stats <- contri_biog_anova %>%
  filter(case_when(name == "response divergence" ~ value != 0, TRUE ~ TRUE)) %>%
  group_by(temp_bin, Lake_biogeographic_status, name) %>%
  summarise(
    mean_val = mean(value, na.rm = TRUE),
    # test full data
    p_value_full = wilcox.test(value, mu = 0, alternative = ifelse(mean_val > 0, "greater", "less"))$p.value,
    p_value_no_out = wilcox.test(
      remove_outliers(value),
      mu = 0,
      alternative = ifelse(mean_val > 0, "greater", "less"),
      na.rm = TRUE
    )$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    sig_full = case_when(
      p_value_full < 0.001 ~ "+++",
      p_value_full < 0.01 ~ "++",
      p_value_full < 0.05 ~ "+",
      TRUE ~ ""
    ),
    sig_no_out = case_when(
      p_value_no_out < 0.001 ~ "(+++)",
      p_value_no_out < 0.01 ~ "(++)",
      p_value_no_out < 0.05 ~ "(+)",
      TRUE ~ ""
    )
  )

# Now calculate y positions per facet group
y_positions <- contri_biog_anova %>%
  group_by(name) %>%
  summarise(
    ymin = min(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    y_pos = ymin * 1,
    y_pos_no_out = y_pos * 1.2
  )

# Join y_positions to stats so each group has the correct y for plotting
stats <- left_join(stats, y_positions, by = c("name"))

#annotation_data <- test_results %>%
#  left_join(label_positions, by = "name") %>%
#  # Convert temp_bin to numeric for positioning
#  mutate(temp_bin_num = as.numeric(factor(temp_bin, levels = unique(temp_bin))))

contri_biog_anova_stats <- left_join(contri_biog_anova, stats)

boxplot_sp_contr <- ggplot(data = contri_biog_anova_stats %>% 
         filter(case_when(name == "response divergence" ~ value != 0, TRUE ~ TRUE)), 
       aes(x = temp_bin, y = value)) +
  
  geom_rect(data = background_rects, inherit.aes = FALSE,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = fill_bg),
            alpha = 0.5) +
  scale_fill_manual(values = c("gray95", "white"), guide = "none") +
  
  geom_hline(aes(yintercept = 0)) +
  
  
  ggbeeswarm::geom_beeswarm(aes(color = Lake_biogeographic_status), 
                            dodge.width = 0.75, cex = 1, priority = "density", 
                            corral.width = 0.5, 
                            alpha = 0.5, 
                            shape = 19, 
                            stroke = 0, 
                            size = 1) +
  
    geom_boxplot(aes(colour = Lake_biogeographic_status), 
               width = 0.5, alpha = 0.3,
               position = position_dodge(width = 0.75), 
               outlier.color = 'gray80', 
               outlier.alpha = 1, 
               outlier.size = 1, 
               outlier.shape = 19, 
               outlier.stroke = 1) +
  
  scale_colour_manual(name = 'biogeography', values = c("#3CB371",  "#C7C35E", 'purple')) +
  
  facet_wrap(~name, scales = "free") +
  
  geom_text(aes(x = temp_bin, group = Lake_biogeographic_status,  colour = Lake_biogeographic_status,
        y = y_pos, label = sig_full ),
    inherit.aes = T,
    size = 3, 
    position = position_dodge(width = 0.75), 
    show.legend = F
  ) +
  geom_text(aes(x = temp_bin, group = Lake_biogeographic_status,  colour = Lake_biogeographic_status,
        y = y_pos_no_out, label = sig_no_out),
    inherit.aes = T,
    size = 3, 
    position = position_dodge(width = 0.75), 
    show.legend = F
  ) +

  theme_bw() +
  theme(panel.grid = element_blank(), 
        legend.position = 'bottom', 
        strip.background = element_blank(), 
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(x = "Temperature strata", y = "Species contribution")

boxplot_sp_contr

my_custom_theme <- function() {
  theme_bw() +
    theme(
      panel.grid = element_blank(),
      legend.position = 'bottom',
      strip.background = element_blank(), 
      aspect.ratio = 0.5,
      strip.text = element_text(size = 7), 
      text = element_text(size = 8)
    )
}

# save figure
png("figures/species_contributions/boxplot_contributions_biogeography.png", res = 600, width = 5000, height = 2500)
boxplot_sp_contr + my_custom_theme()
dev.off()

# save table of tests
contri_biog_anova %>% 
  filter(case_when(name == "response divergence" ~ value != 0, TRUE ~ TRUE)) %>%
  group_by(temp_bin, Lake_biogeographic_status, name) %>%
  summarise(
    mean_val = mean(value, na.rm = TRUE),
    data = list(tibble(value = value, species_dropped = species_dropped)),
    rank_test = list(
      tidy(wilcox.test(value, mu = 0, alternative = ifelse(mean(value, na.rm = TRUE) > 0, "greater", "less"), conf.int = TRUE))
      ),
    rank_test_no_out= list(
      tidy(wilcox.test(remove_outliers(value), mu = 0, alternative = ifelse(mean_val > 0, "greater", "less"),conf.int = TRUE))
      ),
    .groups = "drop"
    ) %>% 
  select(-data) %>% 
  unnest(c(rank_test, rank_test_no_out), names_sep = '_') %>% 
  select(-mean_val, -contains('_method'), -contains('_conf.')) %>% 
  arrange(name, temp_bin, Lake_biogeographic_status) %>% 
  mutate(Lake_biogeographic_status = dplyr::recode(Lake_biogeographic_status, 
                                       'widespread \nnative' = 'WN',
                                       'in situ \ndiversification' = 'ISD',
                                       'non-native' = 'NN'), 
         name = gsub('response ', '', name), 
         temp_bin = gsub('°C', '', temp_bin)) %>% 
  rename('Temperature strata' = temp_bin,
         'Lake-level biogeo. group' = Lake_biogeographic_status, 
         'Response diversity metric' = name) %>% 
  mutate(across(contains('p.value'), signif, 2),
         across(contains('estimate'), signif, 2), 
         across(contains('p.value'),  # replace with your actual p-value columns
                ~ case_when(
                  .x < 0.001 ~ "***",
                  .x < 0.01 ~ "**",
                  .x < 0.05 ~ "*",
                  TRUE ~ ""),
                .names = "sig_{.col}"
                )) %>% 
  relocate('Response diversity metric', 'Temperature strata', 'Lake-level biogeo. group', 
           4,5,6,12,7,8,9,10,13,11,12) %>% 
  write_csv('figures/species_contributions/biogeo_contr_wilcoxon_tests.csv')


```


# Testing variation within species 

## Between lake variation in responses

```{r}

dir.create('figures/species_contributions/', recursive = T)

# process biogeographic data per lake
species_biog_lake <- read_csv("data/taxonomy_May2025/species_lake_list_raw_EDITED.csv") %>% 
  select(Species_ecotype, Lake, Lake_biogeographic_status) %>% 
  unique() %>%
  rename(Species = Species_ecotype) %>% 
  filter(Lake_biogeographic_status != 'remove', 
         !is.na(Lake_biogeographic_status))

# Read in the original response diversity information
response_diversity <- readRDS('data/response_diversity.rds')

# rename 
response_diversity <- response_diversity %>% 
  rename(diss_com = dissimilarity_ross, 
         div_com = divergence_ross, 
         mean_com = mean_ffd, 
         sd_com = sd_ffd)

# Load the data.table containing species contributions
sp_con_rd <- readRDS('data/species_contributions_respdiv.rds')

# get response diversity contributions compared with including species
contributions_full <- left_join(sp_con_rd, response_diversity) %>% 
  filter(!is.na(div_com)) %>% 
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble()

# aggregate as mean for each lake and species
contributions_net <-contributions_full %>% 
  as_tibble() %>% 
  group_by(species_dropped, lake) %>% 
  summarise(across(diss_contr:sd_contr, c(meanLS = mean, sdLS = sd)))


# namving convention for tidier aovs
cn <- c('species' = 'species_dropped', 'lake' = 'lake',  'temperature' = 'temp_bin', 'biogeography' = 'Lake_biogeographic_status')

# add in temperature and biogeography
contributions_biog_temp <-  left_join(sp_con_rd, response_diversity) %>%
  
  # join in biogeographic status at species and lake scale
  left_join(., species_biog_lake %>% unique(), by = c('species_dropped' = 'Species', 'lake' = 'Lake')) %>% 
  
  # remove communities without estimates
  filter(!is.na(div_com), !is.na(Lake_biogeographic_status)) %>% 
  
  # contributions are positive if when you drop the species you lose rd 
  mutate(diss_contr = diss_com - dissimilarity_ross, 
         div_contr = div_com - divergence_ross, 
         mean_contr = mean_com - mean_ffd,
         sd_contr = sd_com - sd_ffd) %>% 
  as_tibble() %>% 
  
  # Create temperature bins 
  mutate(temp_bin = paste0('temp_', sprintf("%02d", plyr::round_any(mean_last_7days, accuracy = 5)))) %>% 
  
  # clean up names for saving aov
  rename(cn) 


# clean names
vn <- c('diss_contr' = 'dissimilarity', 
        'div_contr' = 'divergence', 
        'mean_contr' = 'direction', 
        'sd_contr' = 'SD')

# testing
#var <- 'diss_contr'

# calculate the analysis of variance across all RD responses
aov_sp_contri <- lapply(names(vn), function(var){
  
  var_diff <- setdiff(names(cn), 'species')
  
  fmla <- as.formula(paste(var, "~ species + (", paste(var_diff, collapse = " + "), ")^2"))

  # calculate the mean response in a lake and estimate variaton across lakes
  lm(fmla, 
     data = contributions_biog_temp) %>% 
    aov %>% 
    broom::tidy() %>% 
    mutate(across(c(sumsq, meansq, statistic), signif, digits = 2), 
           p.value = case_when(p.value < 0.001 ~ "<0.001",
                               p.value < 0.01 ~ "<0.01",
                               p.value < 0.05 ~ "<0.05",
                               TRUE ~ as.character(round(p.value, 2)),
                               )) %>% 
    # process sum of squares
    mutate(# Proportion of sum of squares
      prop_sumsq = signif(sumsq / sum(sumsq, na.rm = TRUE), 2),
      # Proportion of mean sum of squares
      prop_meansq = signif(meansq / sum(meansq, na.rm = TRUE), 2), 
      response = vn[names(vn) %in% var]) %>% 
    # related response to first column
    relocate(response) %>% 
    # print output for evaluation in line
    {print(.); .} %>%
    # save output
    {write_csv(., file = paste0('figures/species_contributions/', vn[names(vn) %in% var], '_aov.csv')); .}
    
})

# return objects and save as large table 

bind_rows(aov_sp_contri) %>% 
  relocate(prop_sumsq, .after = sumsq) %>% 
  relocate(prop_meansq, .after = meansq) %>% 
  write_csv(., file = 'figures/species_contributions/all_aov.csv')


# comparing between lakes - what % of species had both positive and negative contributions
contributions_biog_temp %>% 
  group_by(species, lake) %>% 
  summarise(across(contains('_contr'), mean)) %>% 
  ungroup() %>% 
  group_by(species) %>% 
  summarise(across(diss_contr:sd_contr, function(x){if_else(length(unique(na.omit(sign(x)))) == 2, T, F)})) %>% 
  ungroup() %>% 
  summarise(across(contains('_'), function(x) (sum(x) / 52)*100))
spContributions_biog_lake_temp_contributions
# comparing within a lake - what % of species had both positive and negative contributions
contributions_biog_temp %>% 
  group_by(species, lake) %>% 
  summarise(across(diss_contr:sd_contr, function(x){if_else(length(unique(na.omit(sign(x[x!=0])))) == 2, T, F)})) %>% 
  ungroup() %>% 
  group_by(lake) %>% 
  summarise(across(contains('_'), function(x) (sum(x) / length(x))*100)) %>% 
  {print(.); .} %>% 
  ungroup() %>% 
  summarise(across(contains('_'), list(m = mean, sd = sd))) %>% 
  t

```



## Mean variation within a lake (across temperature bands)